---
title: "Homework 2"
author: "Gifty Osei"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{pdfpages}
---

```{r setup, include=FALSE}
<<<<<<< HEAD
=======
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, tidy = TRUE,dev = "cairo_pdf")
>>>>>>> 1bc8b7cb66b86ceda6034867bd57cbe91ba97f98

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, tidy = TRUE,dev = "cairo_pdf")
library(latex2exp)
<<<<<<< HEAD

=======
library(dplyr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(forcats)
library(tidyr)
library(cobalt)
library(tibble)
library(Matching)
library(pander)
library(broom)
library(broom.mixed)
>>>>>>> 1bc8b7cb66b86ceda6034867bd57cbe91ba97f98
```


## Acknowledgement / Disclosure Statement

I discussed this homework with Jacob Antici at some point specifically on NAN's estimation problems, however my work is completely independent. I also used Chat-GPT to fine tune my writings and occasionally in helping to understand with errors and fixing bugs in my code.

<<<<<<< HEAD
=======


## Part 1 Q1a

>>>>>>> 1bc8b7cb66b86ceda6034867bd57cbe91ba97f98
```{r}

## small laptop path
path <- "D:/WashU/PHD/Year 2/Sem 2/SDS 5320/Homework/PS2_data.txt" 

# desktop path
#path <- "F:/PHD/SDS 5320/Homework/PS2_data.txt"
df <- read.table(path, header = TRUE)


# lets look at the freq of the treatment groups and response
#str(df)
kable(table(df$pg))
kable(table(df$i_aqoc))

## physician 1 vs physician 2

```


To define treatment, we convert categorical covariates; Let 

$Z=1$ if $\text{pg} == 1$ (physician group 1) and $Z=0$ if $\text{pg}== 2$

 - i_educ == 1: occurs once and only in one group
 
 - i_insu == 3: occurs once and only in one group
 
 Now to fit the propensity score first we stabilize these covariates;
 
```{r}

# Encoding the binary variables

# convert to 0, 1 treatment, control groups
df$Z <- ifelse(df$pg == 1, 1, 0)

# Convert categorical variables to factors

df$i_sex  <- factor(df$i_sex)
df$i_race <- factor(df$i_race)
df$i_educ <- factor(df$i_educ)
df$i_insu <- factor(df$i_insu)
df$i_drug <- factor(df$i_drug)
df$i_seve <- factor(df$i_seve)

## rename response var
df$outcome <- df$i_aqoc


```
 
```{r}
set.seed(90)
## Fitting the Propensity score model
ps_mod <- glm(
  Z ~ i_age + i_sex + i_race + i_educ + i_insu + i_drug + i_seve +
    com_t + pcs_sd + mcs_sd ,
  data = df,
  family = binomial()
)

# summary
pander(summary(ps_mod), digits = 3, caption = "Coeffiecients From PS model" )

## Predicted propensity score e_hat = P(Z=1 | X)
df$e_hat <- predict(ps_mod, type = "response")

## Predicted weighted IPW 
df$w_ipw <- ifelse(df$Z ==1, 1/df$e_hat, 1/(1-df$e_hat))

## Summary
summary(df$e_hat)
summary(df$w_ipw)

#tapply(df$e_hat, df$Z, summary)

#max(df$w_ipw)

```

Some of the propensity scores are near 0 and 1 so we check weight extremes

We compute IPW (ATE) weights:

 - if Z = 1: $w_1(x_i) = \dfrac{1}{\hat{e}(x_i)}$
 
 - if Z = 0: $w_0(x_i) = \dfrac{1}{(1-\hat{e}(x_i))}$
 
 
```{r}

## IPW weights (ATE)

df$w_ipw <- ifelse(df$Z == 1, 1 / df$e_hat, 1 / (1 - df$e_hat))

## look at the summary
#kable(summary(df$w_ipw), digits = 3, caption = "Summary of Weighted IPW")
#quantile(df$w_ipw, probs = c(.9, .95, .99))

## Show the most extreme weights
df_extreme <- df[order(df$w_ipw, decreasing = TRUE),
                 c("pg","Z","i_aqoc","e_hat","w_ipw","i_race","i_educ","i_insu","i_seve",
                   "com_t","pcs_sd","mcs_sd")]
kable(head(df_extreme,10), caption = "Extreme Weights")

```
 
For Balance checking using ASD, we use;

$$\text{ASD} = \dfrac{|\frac{\sum_{i=1}^N w_1(X_i)Z_i X_{pi}}{\sum_{i=1}^N w_1(X_i)Z_i} - \frac{\sum_{i=1}^N w_0(X_i)(1-Z_i) X_{pi}}{\sum_{i=1}^N w_0(X_i)(1-Z_i)}|}{\sqrt{\frac{s_1^2 + s_0^2}{2}}}$$




For binary covariates, we use the proportions = $(p_1 + p_0)/2$


```{r}

## ASD functions

w_mean <- function(x, w) sum(w * x) / sum(w)
w_var  <- function(x, w) {
  m <- w_mean(x, w)
  sum(w * (x - m)^2) / sum(w)
}

asd_cont <- function(x, z, w1 = NULL, w0 = NULL) {
  if (is.null(w1)) w1 <- rep(1, length(x))
  if (is.null(w0)) w0 <- rep(1, length(x))
  x1 <- x[z == 1]; x0 <- x[z == 0]
  ww1 <- w1[z == 1]; ww0 <- w0[z == 0]
  m1 <- w_mean(x1, ww1); m0 <- w_mean(x0, ww0)
  v1 <- w_var(x1, ww1);  v0 <- w_var(x0, ww0)
  denom <- sqrt((v1 + v0) / 2)
  abs(m1 - m0) / denom
}

asd_bin <- function(xbin, z, w1 = NULL, w0 = NULL) {
  if (is.null(w1)) w1 <- rep(1, length(xbin))
  if (is.null(w0)) w0 <- rep(1, length(xbin))
  x1 <- xbin[z == 1]; x0 <- xbin[z == 0]
  ww1 <- w1[z == 1];  ww0 <- w0[z == 0]
  p1 <- w_mean(x1, ww1); p0 <- w_mean(x0, ww0)
  p <- (p1 + p0) / 2
  denom <- sqrt(p * (1 - p))
  abs(p1 - p0) / denom
}

asd_factor_max <- function(f, z, w1 = NULL, w0 = NULL) {
  levs <- levels(f)
  asds <- sapply(levs, function(L) {
    xbin <- as.integer(f == L)
    asd_bin(xbin, z, w1, w0)
  })
  max(asds, na.rm = TRUE)
}

```


Computing ASD before weighting vs after IPW as;


```{r}
##  ASD table: unweighted vs IPW

z <- df$Z

## group-specific weights for balance computation
w1 <- 1 / df$e_hat
w0 <- 1 / (1 - df$e_hat)

out <- data.frame(
  covariate = c("i_age","com_t","pcs_sd","mcs_sd",
                "i_sex","i_race","i_educ","i_insu","i_drug","i_seve"),
  ASD_unweighted = NA_real_,
  ASD_IPW = NA_real_
)



bal.tab(ps_mod, weights = df$w_ipw, method = "weighting", estimand = "ATE")

```



```{r, eval=FALSE, include=FALSE}
#lets do the love plot here**

library(cobalt)

# Covariate formula (same as your PS model)
f <- Z ~ i_age + i_sex + i_race + i_educ + i_insu + i_drug + i_seve + com_t + pcs_sd + mcs_sd

# Balance table for IPW (includes BOTH unadjusted + adjusted automatically)
bal_ipw <- bal.tab(
  f,
  data = df,
  weights = df$w_ipw,
  method = "weighting",
  estimand = "ATE",
  s.d.denom = "pooled"
)

print(bal_ipw)

# # Love plot: Unweighted vs IPW
# love.plot(
#   bal_ipw,
#   stat = "mean.diffs",     # ASD / standardized mean differences
#   abs  = TRUE,
#   thresholds = c(m = .1),  # ASD=0.1 reference line
#   var.order = "unadjusted",
#   line = TRUE
# )


bal_ow <- bal.tab(
  f,
  data = df,
  weights = df$w_ow,
  method = "weighting",
  estimand = "ATO",
  s.d.denom = "pooled"
)

print(bal_ow)
# 
# love.plot(
#   bal_ow,
#   stat = "mean.diffs",
#   abs  = TRUE,
#   thresholds = c(m = .1),
#   var.order = "unadjusted",
#   line = TRUE
# )


## Make sure treatment exists in df
df$Z <- ifelse(df$pg == 1, 1, 0)
table(df$Z)

## Balance object (IPW)
f <- Z ~ i_age + i_sex + i_race + i_educ + i_insu + i_drug + i_seve + com_t + pcs_sd + mcs_sd

bal_ipw <- bal.tab(
  f,
  data = df,
  weights = df$w_ipw,
  method = "weighting",
  estimand = "ATE",
  s.d.denom = "pooled"
)

## Love plot (USE bal_ipw, not df)
# love.plot(
#   bal_ipw,
#   stat = "mean.diffs",
#   abs = TRUE,
#   thresholds = c(m = 0.1),
#   var.order = "unadjusted",
#   line = TRUE
# )


X <- model.matrix(
  ~ i_age + i_sex + i_race + i_educ + i_insu + i_drug + i_seve + com_t + pcs_sd + mcs_sd,
  data = df
)[, -1, drop = FALSE]

bal_ipw2 <- bal.tab(
  X,
  treat = df$Z,
  weights = df$w_ipw,
  method = "weighting",
  estimand = "ATE",
  s.d.denom = "pooled"
)

# love.plot(
#   bal_ipw2,
#   stat = "mean.diffs",
#   abs = TRUE,
#   thresholds = c(m = 0.1),
#   var.order = "unadjusted",
#   line = TRUE
# )

love.plot(
  list("IPW (ATE)" = bal_ipw, "OW (ATO)" = bal_ow),
  stat = "mean.diffs",
  abs  = TRUE,
  thresholds = c(m = .1),
  var.order = "unadjusted",
  line = TRUE
)

```

**Comments**

We see that after applying the weights, nearly all covariates are at a good balance using the ASD $\leq$ 0.1. We see that IPW drastically improves balance for most covariates. Variable 'Race' still remains poorly balanced and this is because 1 race category has very limited overlap across physician groups. We cannot track any evidence of extreme IP weights given the threshold.


## Part 1 Q1b

```{r}

## PS model  again
ps_mod <- glm(
  Z ~ i_age + i_sex + i_race + i_educ + i_insu + i_drug + i_seve + com_t + pcs_sd + mcs_sd,
  data = df, family = binomial()
)

## store ps values
df$e_hat <- predict(ps_mod, type = "response")

## Weights
df$w_ipw <- ifelse(df$Z==1, 1/df$e_hat, 1/(1-df$e_hat))      # ATE / IPW
df$w_ow  <- ifelse(df$Z==1, 1-df$e_hat, df$e_hat)           # OW / overlap (ATO)


pander(summary(df$w_ow), caption = "Summary for Overlap Weights")

```




```{r, fig.cap="This is a plot of propensity score distribution for the physician group.We see somewhat poor/weak overlap PS meaning control group has many PS near 0 while treatment group has few, which also creates extreme IPW weights."}

df %>%
  mutate(group = factor(Z, levels = c(0,1), labels = c("pg=2 (Z=0)", "pg=1 (Z=1)"))) %>%
  ggplot(aes(x = e_hat, fill = group)) +
  geom_histogram(position = "identity", alpha = 0.45, bins = 30) +
  labs(x = "Estimated propensity score e(X)=P(Z=1|X)", y = "Count", fill = "Group",
       title = "Propensity score distributions by physician group") +
  theme_bw()

```




```{r, fig.cap="The figure shows the love plots for the covariates. It shows that weighting improves the covariate balance where almost all the ASDs are below 0.1." }

w_mean <- function(x, w) sum(w*x) / sum(w)
w_var  <- function(x, w) {
  m <- w_mean(x, w)
  sum(w*(x-m)^2) / sum(w)
}

asd_w <- function(x, z, w) {
  x1 <- x[z==1]; x0 <- x[z==0]
  w1 <- w[z==1]; w0 <- w[z==0]
  m1 <- w_mean(x1, w1); m0 <- w_mean(x0, w0)
  v1 <- w_var(x1, w1);  v0 <- w_var(x0, w0)
  denom <- sqrt((v1+v0)/2)
  abs(m1-m0)/denom
}

## Build design matrix (no intercept) so factors become dummies
X <- model.matrix(
  ~ i_age + i_sex + i_race + i_educ + i_insu + i_drug + i_seve + com_t + pcs_sd + mcs_sd,
  data = df
)[, -1, drop = FALSE]  # drop intercept

asd_table <- function(w, label) {
  tibble(
    covariate = colnames(X),
    ASD = apply(X, 2, function(x) asd_w(x, df$Z, w)),
    scheme = label
  )
}

asds <- bind_rows(
  asd_table(rep(1, nrow(df)), "Unweighted"),
  asd_table(df$w_ipw, "IPW (ATE)"),
  asd_table(df$w_ow,  "OW (ATO)")
)

## Love plot using ggplot
# asds %>%
#   mutate(scheme = factor(scheme, levels = c("Unweighted","IPW (ATE)","OW (ATO)")),
#          covariate = fct_reorder(covariate, ASD, max)) %>%
#   ggplot(aes(x = ASD, y = covariate, shape = scheme)) +
#   geom_point(size = 2.3) +
#   geom_vline(xintercept = 0.1, linetype = "dashed") +
#   labs(x = "Absolute Standardized Difference (ASD)", y = NULL, shape = NULL,
#        title = "Love plot: balance before and after weighting") +
#   theme_bw()

love.plot(ps_mod, weights = list(IPW = df$w_ipw, OW = df$w_ow), stats = "mean.diffs",
abs = TRUE, thresholds = 0.1, sample.names = c("Unweighted", "IPW","OW"), 
title = "Covariate Balance by Love Plot",

var.order = "unadjusted", stars = "std")

```




## Part 1 Q1c

Overlap weighting (OW) which balances model covariates automatically

We now; $e(X) = P(Z =1 \mid X)$

 - If Z = 1: $w^{OW} = 1 - e(X)$
 
 - If Z = 0: $w^{OW} = e(X)$

```{r}
## Overlap weighting

w_group_mean <- function(x, z, w, group_val) {
  idx <- (z == group_val)
  w_mean(x[idx], w[idx])
}

means_table <- function(w, label) {
  tibble(
    covariate = colnames(X),
    mean_Z1 = apply(X, 2, function(x) w_group_mean(x, df$Z, w, 1)),
    mean_Z0 = apply(X, 2, function(x) w_group_mean(x, df$Z, w, 0)),
    scheme = label
  )
}

tab_means <- bind_rows(
  means_table(rep(1, nrow(df)), "Unweighted"),
  means_table(df$w_ipw,         "IPW (ATE)"),
  means_table(df$w_ow,          "OW (ATO)")
) %>%
  mutate(across(c(mean_Z1, mean_Z0), ~round(.x, 3))) %>%
  arrange(covariate, scheme)

pander(tab_means)


tab_means_wide <- tab_means %>%
  pivot_wider(
    names_from = scheme,
    values_from = c(mean_Z1, mean_Z0)
  )

pander(tab_means_wide)


```

\newpage

```{r}


## Build a covariate matrix that includes ALL factor levels
X_cat <- model.matrix(
  ~ 0 + i_sex + i_race + i_educ + i_insu + i_drug + i_seve,
  data = df
)

X <- cbind(
  i_age  = df$i_age,
  com_t  = df$com_t,
  pcs_sd = df$pcs_sd,
  mcs_sd = df$mcs_sd,
  X_cat
)

z     <- df$Z
w_ipw <- df$w_ipw
w_ow  <- df$w_ow

wmean <- function(x, w) sum(w * x) / sum(w)

## Compute unadjusted group means, overall, and weighted adjusted means 
tab1 <- tibble(
  
  covariate = colnames(X),
  
  `Z=1`     = colMeans(X[z == 1, , drop = FALSE]),
  
  `Z=0`     = colMeans(X[z == 0, , drop = FALSE]),
  
  `Overall` = colMeans(X),
  
  `IPW`     = apply(X, 2, wmean, w = w_ipw),
  
  `OW`      = apply(X, 2, wmean, w = w_ow)
) %>%
  mutate(across(-covariate, ~ round(.x, 3)))

##  other
tab1 <- tab1 %>%
  mutate(
    covariate = gsub("^i_([a-z]+)(.*)$", "i_\\1\\2", covariate),
    covariate = gsub("(i_sex|i_race|i_educ|i_insu|i_drug|i_seve)(.+)",
                     "\\1 =\\2", covariate)
  )

##  creating the table in lecture notes
N1 <- sum(z == 1)
N0 <- sum(z == 0)
N <- length(z)

colnames(tab1) <- c(
  "Covariate",
  paste0("Group 1 (Z=1)\nN=", N1),
  paste0("Group 2 (Z=0)\nN=", N0),
  paste0("Overall\nN=", N),
  "IPW weighted",
  "Overlap weighted"
)

kable(tab1,
      caption = "Table 1: Covariate means (unadjusted and weighted)") %>%
  add_header_above(c(" " = 1, "Unadjusted" = 3, "Adjusted" = 2)) %>%
  kable_styling(full_width = FALSE)

```



## Part 1 Q1d

To find ATE(IPW) and ATO(OW) using HT (non normalized) and Hajek (normalized)

Let h(X) define the tilting function as the target population:

 - IPW(ATE): h(X) = 1
 
 - OW (ATO): h(X) = e(X){1-e(X)}
 
and weights are;

 - $w_1(X) = \dfrac{h(X)}{e(X)}$ for the treated (Z = 1)
 
 - $w_0(X) = \dfrac{h(X)}{(1-e(X))}$ for the control (Z = 0)
 
 
The estimators are;


 - Non-normalized (HT)
 
  $$\hat{\tau}_{h,1} = \dfrac{\sum w_1 ZY}{\sum h} - \dfrac{\sum w_0 (1-Z)Y}{\sum h}$$
  
  
  
  
  
 - Normalized (Hajek)
 
    $$\hat{\tau}_{h,2} = \dfrac{\sum w_1 ZY}{\sum w_1 Z} - \dfrac{\sum w_0 (1-Z)Y}{\sum w_0 (1-Z)}$$
    
    
```{r}

set.seed(45)
## Weighted ATE - WATE

tau_hat <- function(y, z, e, h) {
  w1 <- h / e
  w0 <- h / (1 - e)

   
    denom <- sum(h)
    mu1 <- sum(w1 * z * y)
    mu0 <- sum(w0 * (1 - z) * y)
    tau1 <- (mu1 - mu0) / denom
  
    denom1 <- sum(w1 * z)
    denom0 <- sum(w0 * (1 - z))
    
    mu1 <- sum(w1 * z * y) / denom1
    mu0 <- sum(w0 * (1 - z) * y) / denom0
    tau2 <- mu1 - mu0
    
    c(tau1 =tau1, tau2 = tau2)
}





y <- df$i_aqoc   
z <- as.factor(df$Z)
e <- df$e_hat




## ATE via IPW: h=1
h_ate <- rep(1, length(e))

## ATO via OW: h=e(1-e)
h_ato <- e * (1 - e)



y <- as.numeric(df$outcome)     
z <- as.integer(df$Z == 1)      
e <- as.numeric(df$e_hat)

eps <- 1e-6
e <- pmin(pmax(e, eps), 1 - eps)

h_ate <- rep(1, length(e))
h_ato <- e * (1 - e)



estimates <- tibble(
  estimand = c("ATE (IPW)","ATE (IPW)","ATO (OW)","ATO (OW)"),
  version  = c("HT","Hajek","HT","Hajek"),
  tau      = c(
    tau_hat(y,z,e,h_ate)["tau1"],
    tau_hat(y,z,e,h_ate)["tau2"],
    tau_hat(y,z,e,h_ato)["tau1"],
    tau_hat(y,z,e,h_ato)["tau2"]
  )
)

estimates




```

**Bootstrap SE**

```{r}

set.seed(456)

B <- 1000

n <- length(y)


bootstrap_fun <- replicate(B, {
  idx <- sample.int(n,n, replace = T)
  y_b <- y[idx]
  z_b <- z[idx]
  e_b <- e[idx]
  
  ipw_b <- tau_hat(y_b,z_b,e_b,h = rep(1, length(e_b)))
  
  ow_b <- tau_hat(y_b,z_b,e_b, h = e_b * (1- e_b))
  
 
  
  c(IPW_tau1 = ipw_b["tau1"],IPW_tau2 =ipw_b["tau2"], OW_tau1 = ow_b["tau1"], OW_tau2 = ow_b["tau2"])
})


## apply bootstrap function
bootstrap_se <- apply(bootstrap_fun, 1, sd)




results <- tibble(Estimator = c("ATE (IPW)","ATE (IPW)","ATO (OW)","ATO (OW)"), SE = as.numeric(bootstrap_se))



kable(results)

```





```{r, eval=FALSE, include=FALSE}


## ---- tau_hat (yours) ----
tau_hat <- function(y, z, e, h) {

  idx <- is.finite(y) & is.finite(z) & is.finite(e) & is.finite(h)
  y <- y[idx]; z <- z[idx]; e <- e[idx]; h <- h[idx]

  w1 <- h / e
  w0 <- h / (1 - e)

  # HT (non-normalized)
  denom <- sum(h)
  mu1 <- sum(w1 * z * y)
  mu0 <- sum(w0 * (1 - z) * y)
  tau1 <- (mu1 - mu0) / denom

  # Hajek (normalized)
  denom1 <- sum(w1 * z)
  denom0 <- sum(w0 * (1 - z))
  mu1 <- sum(w1 * z * y) / denom1
  mu0 <- sum(w0 * (1 - z) * y) / denom0
  tau2 <- mu1 - mu0

  c(tau1 = tau1, tau2 = tau2)
}

## ---- inputs (make sure z is 0/1 numeric!) ----
y <- as.numeric(df$i_aqoc)           # or df$outcome
z <- as.integer(df$Z == 1)           # MUST be 0/1
e <- as.numeric(df$e_hat)

eps <- 1e-6
e <- pmin(pmax(e, eps), 1 - eps)

h_ate <- rep(1, length(e))
h_ato <- e * (1 - e)

## ---- point estimates (4 numbers in a named vector) ----
pt <- c(
  IPW_tau1 = tau_hat(y, z, e, h_ate)["tau1"],
  IPW_tau2 = tau_hat(y, z, e, h_ate)["tau2"],
  OW_tau1  = tau_hat(y, z, e, h_ato)["tau1"],
  OW_tau2  = tau_hat(y, z, e, h_ato)["tau2"]
)

## ---- bootstrap SEs (replicate; returns matrix 4 x B) ----
set.seed(123)
B <- 1000
n <- length(y)

boot_mat <- replicate(B, {
  idx <- sample.int(n, n, replace = TRUE)
  y_b <- y[idx]; z_b <- z[idx]; e_b <- e[idx]
  e_b <- pmin(pmax(e_b, eps), 1 - eps)

  ipw_b <- tau_hat(y_b, z_b, e_b, h = rep(1, length(e_b)))
  ow_b  <- tau_hat(y_b, z_b, e_b, h = e_b * (1 - e_b))

  c(
    IPW_tau1 = ipw_b["tau1"],
    IPW_tau2 = ipw_b["tau2"],
    OW_tau1  = ow_b["tau1"],
    OW_tau2  = ow_b["tau2"]
  )
})

boot_se <- apply(boot_mat, 1, sd, na.rm = TRUE)

## ---- final combined table (est + SE) ----
final_tab <- tibble(
  Estimand  = c("ATE (IPW)", "ATE (IPW)", "ATO (OW)", "ATO (OW)"),
  Version   = c("HT", "Hajek", "HT", "Hajek"),
  Estimate  = as.numeric(pt[c("IPW_tau1","IPW_tau2","OW_tau1","OW_tau2")]),
  Boot_SE   = as.numeric(boot_se[c("IPW_tau1","IPW_tau2","OW_tau1","OW_tau2")])
) %>%
  mutate(across(c(Estimate, Boot_SE), ~round(.x, 4)))

kable(final_tab, caption = "Weighted ATE/ATO estimates with bootstrap standard errors") %>%
  kable_styling(full_width = FALSE)
```


**Comments**

IPW (ATE) vs OW (ATO)

 -  Different estimands: IPW targets the ATE over the full covariate distribution (full population); OW targets the ATO over the overlap population where both groups are plausible.
 
 - Variance/stability: IPW can have large variance if some $\hat{e}(X)$'s are near 0/1 which implies huge weights. OW downweights extremes and is usually more stable with better balance.
 
Under the Normalized Hajek vs Non-normalized HT within a scheme;

 - HT uses $\sum h(X_i)$ in the denominator (stays the same for both groups); it is more sensitive to weight variability.
 
 - Hajek normalizes within treatment groups; it often has smaller variance and better finite-sample behavior.
 
 - The can differ noticeably when weights are highly variable or overlap is weak.
 
 
 
  
## Part 1 Q2

The estimand we want is 

$$E[Y(1) | V = 1] - E[Y(2)|V  = 1]$$

```{r}
set.seed(67)

idx <- as.character(df$i_sex) == "1"

cat("\n counts of treated in V= 1: \n")

print(table(df$Z[idx]))


Z1 <- df$Z[idx]
Y1 <- df$outcome[idx]
e <- df$e_hat[idx]


## IPW estimator for question 2
ipw_2 <- function(Z, Y, e) {
  eps <- 1e-6
  e <- pmin(pmax(e, eps), 1- eps)
  w1 <- 1/e
  w0 <- 1/(1-e)
  
  tau1 <- mean(w1 * Z * Y) - mean(w0 * (1 - Z) * Y)
  tau2 <- sum(w1 * Z * Y)/sum(w1 * Z) - sum(w0 * (1 - Z) * Y)/sum(w0 * (1 - Z))
  
  c(tau1 = tau1, tau2 = tau2)
}

```

2a. We fit PS using all data including V, then do IPW within V = 1. Reason being the PS model be $e(X,V)= p(z =1|x,V)$. Then restricting to V = 1 and weighing by $\dfrac{1}{e(X)}$ and $\dfrac{1}{(1-e(X))}$ identifies the subgroup causal effect.

```{r}
a_ps <-  glm(Z ~ i_age + i_sex + i_race + i_educ + i_insu + i_drug + i_seve + com_t + pcs_sd + mcs_sd,
  data = df, family = binomial())

e_a <- predict(a_ps, type = "response")
a_est <- ipw_2(Z1, Y1, e_a[idx])
```


2b. Fitting PS excluding V, then do IPW within V = 1

In general it is wrong unless $Z \perp V \mid X$ (i.e V adds no information about treatment given X). Reason: If treatment depends on V beyond X, then using e(X) instead of e(X,V) does not balance V-related treatment selection inside the subgroup.

```{r}
b_ps <-  glm(Z ~ i_age + i_race + i_educ + i_insu + i_drug + i_seve + com_t + pcs_sd + mcs_sd,
  data = df, family = binomial())

e_b <- predict(b_ps, type = "response")

b_est <- ipw_2(Z1, Y1, e_b[idx])
```


2c. Restrict to V = 1, the estimates PS within V = 1 excluding V then IPW: This is correct because within subgroup, the correct PS is $e_{V=1} (X) =  P(Z = 1\mid X, V = 1)$. Estimating it using only V = 1 data is fine (often better) and then IPW targets that subgroup

```{r}
c_ps <-  glm(Z ~ i_age + i_race + i_educ + i_insu + i_drug + i_seve + com_t + pcs_sd + mcs_sd,
  data = df[idx,], family = binomial())

e_c <- predict(c_ps, type = "response")
c_est <- ipw_2(Z1, Y1, e_c)
```


```{r}

comp <- data.frame(
Method = c("(a) PS includes V; IPW in V=1",
"(b) PS excludes V; IPW in V=1",
"(c) PS fit within V=1 (excl V); IPW in V=1"),
tau1 = c(a_est["tau1"], b_est["tau1"], c_est["tau1"]),
tau2 = c(a_est["tau2"], b_est["tau2"], c_est["tau2"])
)
comp$tau1 <- round(comp$tau1, 4)
comp$tau2 <- round(comp$tau2, 4)


print(comp, row.names = FALSE)
```


 **Bootstrap SE**
 
```{r}
boot_se_fixed_e <- function(Y, Z, e, B = 1000){
n <- length(Y)
boots <- replicate(B, {
ii <- sample.int(n, n, replace = TRUE)
ipw_2(Z[ii], Y[ii], e[ii])
})
apply(boots, 1, sd)
}

se_a <- boot_se_fixed_e(Y1, Z1, e_a[idx], B = 1000)
se_c <- boot_se_fixed_e(Y1, Z1, e_c, B = 1000)
cat("\nBootstrap SEs (correct methods):\n")

cat("Method (a): SE(tau1) =", round(se_a["tau1"], 4), " SE(tau2) =",
round(se_a["tau2"], 4), "\n")

cat("Method (c): SE(tau1) =", round(se_c["tau1"], 4), " SE(tau2) =",
round(se_c["tau2"], 4), "\n")
```
 
 
```{r, eval=FALSE, include=FALSE}
boot_q2 <- function(df, B=500, seed=123, eps=1e-6) {
  set.seed(seed)
  n <- nrow(df)
  out <- matrix(NA_real_, nrow=B, ncol=3)
  colnames(out) <- c("a","b","c")

  b <- 1
  while (b <= B) {
    idx <- sample.int(n, n, replace=TRUE)
    d <- df[idx, , drop=FALSE]

    ## restore factor levels
    for (v in names(global_levels)) {
      d[[v]] <- factor(d[[v]], levels = global_levels[[v]])
    }

    ## need both treatment groups INSIDE V=1
    dV1 <- d[d$V1==1, , drop=FALSE]
    if (nrow(dV1) < 10) next
    if (length(unique(dV1$Z)) < 2) next

    est <- tryCatch(fit_and_estimate(d), error=function(e) NULL)
    if (is.null(est)) next

    ## bound checks can be added inside fit_and_estimate if needed
    out[b,] <- est
    b <- b + 1
  }

  se <- apply(out, 2, sd, na.rm=TRUE)
  list(reps=out, se=se)
}

bs2 <- boot_q2(df, B=500, seed=123)
bs2$se

## Summary table
res_q2 <- data.frame(
  method = c("(a) PS includes V; weight within V=1",
             "(b) PS excludes V; weight within V=1",
             "(c) Fit PS within V=1 (exclude V)"),
  estimate = unname(point_est),
  boot_se  = c(bs2$se["a"], bs2$se["b"], bs2$se["c"])
)
res_q2

```

 **Comments**
 
 The correct method are a and c; a is correct because it uses the correct PS e(X,V) and then targets the subgroup V =1. C is correct because within the subgroup V = 1, the relevant PS is $P(Z = 1 \mid X,V = 1)$, which is exactly what c estimates. b is not generally correct because it uses e(X) rather than e(X,V); if treatment assignment differs by sex beyond other covariates, it fails to balance the selection mechanism for the subgroup effect.
 
 
## Part 1 Q3
 
Outcome here is Y = i_aqoc (1 satisfied, 0 not). The target estimand is the marginal casual odds ratio. 
 $$Q = \dfrac{p_1 (1-p_1)}{p_2 (1-p_2)}$$ , $$p_z = Pr (Y(z) = 1)$$
 

```{r, warning=FALSE, message=FALSE}

# re load data
df <- read.table("D:/WashU/PHD/Year 2/Sem 2/SDS 5320/Homework/PS2_data.txt", header = TRUE)

y  <- df$i_aqoc
z1 <- as.integer(df$pg == 1)   

# X covariates
Xvars <- c("i_age","i_sex","i_race","i_educ","i_insu","i_drug","i_seve",
           "com_t","pcs_sd","mcs_sd")

dat <- df[, c("i_aqoc","pg",Xvars)]
dat$y  <- y
dat$z1 <- z1
dat <- dat[complete.cases(dat[, c("y","z1",Xvars)]), ]

# set as character
for(v in Xvars){
  if(is.character(dat[[v]])) dat[[v]] <- factor(dat[[v]])
}
for(v in Xvars){
  if(is.factor(dat[[v]])) dat[[v]] <- factor(dat[[v]], levels = levels(dat[[v]]))
}

or_from_p <- function(p1,p2) (p1*(1-p2)) / (p2*(1-p1))


# Outcome model: y ~ z1 * X  but fitted on a full-rank matrix

fit_outcome <- function(d){
  # build model matrix with fixed factor levels
  f <- as.formula(paste0("~ z1 * (", paste(Xvars, collapse=" + "), ")"))
  X <- model.matrix(f, data=d)
  
  keep <- qr(X)$pivot[seq_len(qr(X)$rank)]
  Xr <- X[, keep, drop=FALSE]

  glm.fit(x = Xr, y = d$y, family = binomial())
}

## prediction function
pred_outcome <- function(fit, d, zval){
  f <- as.formula(paste0("~ z1 * (", paste(Xvars, collapse=" + "), ")"))
  d2 <- d
  d2$z1 <- zval
  X <- model.matrix(f, data=d2)
  # match columns used in fit
  X <- X[, colnames(fit$xlevels %||% fit$coefficients) , drop=FALSE] 
}

# safer predict 
predict_from_fit <- function(fit, d, zval){
  f <- as.formula(paste0("~ z1 * (", paste(Xvars, collapse=" + "), ")"))
  d2 <- d; d2$z1 <- zval
  X <- model.matrix(f, data=d2)

  # fit using glm.fit 
  Xr <- X[, fit$Xcols, drop=FALSE]
  as.numeric(plogis(Xr %*% fit$coefficients))
}

## the other option fit function
fit_outcome2 <- function(d){
  f <- as.formula(paste0("~ z1 * (", paste(Xvars, collapse=" + "), ")"))
  X <- model.matrix(f, data=d)

  keep <- qr(X)$pivot[seq_len(qr(X)$rank)]
  Xr <- X[, keep, drop=FALSE]

  
  Xr <- as.matrix(Xr)
  storage.mode(Xr) <- "double"

  fit <- glm.fit(x = Xr, y = d$y, family = binomial())

  fit$Xcols <- colnames(Xr)
  fit$Xmat  <- Xr         
  fit
}


```

## Part 1 Q3a

```{r}
## estimation
est_a <- function(d){
  fit <- fit_outcome2(d)
  if(!("z1" %in% names(fit$coefficients))) return(c(est=NA_real_, se=NA_real_))

  delta <- fit$coefficients["z1"]

  X  <- fit$Xmat
  mu <- as.numeric(fit$fitted.values)
  W  <- mu * (1 - mu)

  # X' W X where W is diagonal =
  Xw <- X * sqrt(W)
  XtWX <- crossprod(Xw)

  V <- tryCatch(solve(XtWX), error=function(e) NULL)
  if(is.null(V)) return(c(est=exp(delta), se=NA_real_))

  j <- which(colnames(X) == "z1")
  se_delta <- sqrt(V[j, j])

  c(est = exp(delta), se = exp(delta) * se_delta)
}


```

## Part 1 Q3b

```{r}
# average f(x,z) over X in that group
est_b <- function(d){
  fit <- fit_outcome2(d)
  if(any(is.na(fit$coefficients))) return(c(est=NA_real_))
  p1 <- mean(predict_from_fit(fit, d[d$z1==1, , drop=FALSE], 1))
  p2 <- mean(predict_from_fit(fit, d[d$z1==0, , drop=FALSE], 0))
  c(est = or_from_p(p1,p2))
}


```


## Part 1 Q3c

**PS Stratification**

```{r}
# average f(x,z) over ALL X
est_c <- function(d){
  fit <- fit_outcome2(d)
  if(any(is.na(fit$coefficients))) return(c(est=NA_real_))
  p1 <- mean(predict_from_fit(fit, d, 1))
  p2 <- mean(predict_from_fit(fit, d, 0))
  c(est = or_from_p(p1,p2))
}


# Propensity score strata (5 blocks)

fit_ps <- function(d){
  f <- as.formula(paste0("~ ", paste(Xvars, collapse=" + ")))
  X <- model.matrix(f, data=d)
  keep <- qr(X)$pivot[seq_len(qr(X)$rank)]
  Xr <- X[, keep, drop=FALSE]
  fit <- glm.fit(x=Xr, y=d$z1, family=binomial())
  fit$Xcols <- colnames(Xr)
  list(fit=fit, f=f)
}

predict_ps <- function(ps, d){
  X <- model.matrix(ps$f, data=d)
  Xr <- X[, ps$fit$Xcols, drop=FALSE]
  as.numeric(plogis(Xr %*% ps$fit$coefficients))
}

make_strata5 <- function(d){
  ps <- fit_ps(d)
  e  <- predict_ps(ps, d)
  cuts <- unique(as.numeric(quantile(e, probs=seq(0,1,0.2), na.rm=TRUE)))
  if(length(cuts) < 3) return(rep(1L, nrow(d))) # degenerate case
  cut(e, breaks=cuts, include.lowest=TRUE, labels=FALSE)
}


```

## Part 1 Q3d

```{r}
# compute Odds Ratio within each stratum and take average 
est_d <- function(d){
  S <- make_strata5(d)
  ors <- c()
  for(s in sort(unique(S))){
    ds <- d[S==s, ]
    if(length(unique(ds$z1)) < 2) next
    a <- sum(ds$y==1 & ds$z1==1)
    b <- sum(ds$y==0 & ds$z1==1)
    c0<- sum(ds$y==1 & ds$z1==0)
    d0<- sum(ds$y==0 & ds$z1==0)
    if(any(c(a,b,c0,d0)==0)){ a<-a+0.5; b<-b+0.5; c0<-c0+0.5; d0<-d0+0.5 }
    ors <- c(ors, (a*d0)/(b*c0))
  }
  if(length(ors)==0) return(c(est=NA_real_))
  c(est = mean(ors))
}


```

## Part 1 Q3e
```{r}
# within each stratum, we compute p1s, p2s and take block averages
# to get OR from p1,p2

est_e <- function(d){
  S <- make_strata5(d)
  p1s <- p2s <- c()
  for(s in sort(unique(S))){
    ds <- d[S==s, ]
    if(length(unique(ds$z1)) < 2) next
    p1s <- c(p1s, mean(ds$y[ds$z1==1]))
    p2s <- c(p2s, mean(ds$y[ds$z1==0]))
  }
  if(length(p1s)==0) return(c(est=NA_real_))
  p1 <- mean(p1s); p2 <- mean(p2s)
  c(est = or_from_p(p1,p2))
}


#  bootstrap 

boot_se <- function(d, est_fun, B=500, seed=1){
  set.seed(seed)
  n <- nrow(d)
  vals <- replicate(B, {
    idx <- sample.int(n, n, replace=TRUE)
    out <- tryCatch(est_fun(d[idx, , drop=FALSE]), error=function(e) c(est=NA_real_))
    as.numeric(out["est"])
  })
  sd(vals, na.rm=TRUE)
}


# Compute results for a,b,c,d and e

A <- est_a(dat)
B <- c(est_b(dat), se=boot_se(dat, est_b))
C <- c(est_c(dat), se=boot_se(dat, est_c))
D <- c(est_d(dat), se=boot_se(dat, est_d))
E <- c(est_e(dat), se=boot_se(dat, est_e))

out <- data.frame(
  method = c("a","b","c","d","e"),
  OR_hat = c(A["est"], B["est"], C["est"], D["est"], E["est"]),
  SE     = c(A["se"],  B["se"],  C["se"],  D["se"],  E["se"])
)
kable(out, caption = "Estimates for 3a - 3e")
```


 **Comment**
 
 (c) is the correct one among (a)-(c):
 
 - It computes $p_1 = E[f(X,1)]$ and $p_2 = E[f(X,0)]$ over the same target covariate distribution (overall sample), then forms the OR.
 
 - Under the assumed correct model, this is standard g-computation/standardization for a marginal estimand.
 
 
(e) is also correct as a stratified standardization approach (approximation)

 - Within each PS block, estimate $p_{1k}, p_{2k}$, then average across blocks (weighted by block sizes) to get $p_1, p_2$. then compute OR.
 
 - With sufficiently fine stratification, this approximates the same standardization target.
 
 
These are not generally correct

 (a) is not the marginal OR $Q$
 
 - $exp(\delta)$ is a conditional odds ratio (and with interactions the conditional OR varies with X anyways).
 
 - Odds ratios are non-collapsible, so conditional OR $\ne$ marginal OR in general.
 
(b) is not the target because it averages over different covariate distribution (group 1's X distribution for p1 and group 2's X distribution for p2). That does not correspond to $p_1 = Pr(Y(1) = 1)$ and $p_2 = Pr(Y(1) = 1)$ on a common target population.


(d) is not correct because "average of stratum OR's is not valid aggregation for a marginal OR. If you want a stratification-based OR, the classical approach is Mantel-Haenszel OR, not a simple mean.



## Part 1 Q4

\includepdf[pages=1-3]{D:/WashU/PHD/Year 2/Sem 2/SDS 5320/Homework/Homework2gifty_causal.pdf}


## Part 2

```{r}
# Treatment
df$Z <- as.integer(df$pg == 1)

# Outcome
df$Y <- df$i_aqoc

# Pre-treatment (X covariates)
Xvars <- c("i_age","i_sex","i_race","i_educ","i_insu","i_drug","i_seve",
           "com_t","pcs_sd","mcs_sd")

dat <- df[, c("Y","Z",Xvars)]
dat <- dat[complete.cases(dat), ]


for(v in Xvars){
  if(is.character(dat[[v]])) dat[[v]] <- factor(dat[[v]])
  if(is.factor(dat[[v]])) dat[[v]] <- factor(dat[[v]], levels=levels(dat[[v]]))
}

# Turn X into numeric matrix 
Xmm <- model.matrix(as.formula(paste0("~ ", paste(Xvars, collapse=" + "))), data=dat)
Xmm <- Xmm[, -1, drop=FALSE]  # drop intercept
Xmm <- as.matrix(Xmm); storage.mode(Xmm) <- "double"

Y <- dat$Y
Z <- dat$Z
N <- nrow(dat)
```


## Part 2 Q1a

\includepdf[pages=4-7]{D:/WashU/PHD/Year 2/Sem 2/SDS 5320/Homework/Homework2gifty_causal.pdf}

```{r, eval=FALSE, include=FALSE}
# 1(a) Tuning parameters in matching

cat("\n1(a) Tuning choices in matching (>=4):\n")
cat("  (i) Number of neighbors M: larger M -> lower variance, higher bias (worse matches), more compute.\n")
cat(" (ii) Replacement (TRUE/FALSE): replacement reduces bias (better matches), can increase variance via high reuse; faster/stable.\n")
cat("(iii) Caliper / tolerance (on distance or PS): tighter caliper reduces bias, may increase variance and drop units (less overlap).\n")
cat(" (iv) Distance metric / scaling (Euclidean vs Mahalanobis; standardization): affects bias (quality of matches) and compute.\n")
cat("  (v) Exact / coarsened exact matching on key covariates: reduces bias but may discard many units (variance increases).\n")
cat(" (vi) Common support trimming: reduces extrapolation bias; reduces sample size -> higher variance.\n\n")
```


## Part 2 Q2

```{r}


# normalized Euclidean distance 
# Lets define s_p = SD of (X_ip - X_i'p) across all pairs (i,i').
# For a covariate x, Var(x_i - x_j) = 2 Var(x) over all unordered pairs,
# so SD(diff) = sqrt(2) * sd(x). 
scaled_X_for_covmatch <- function(X){
  s <- apply(X, 2, sd)
  s_diff <- sqrt(2) * s
  s_diff[s_diff == 0] <- 1  
  Xs <- sweep(X, 2, s_diff, "/")
  list(Xs = Xs, s_diff = s_diff)
}

# distance matrix between treated and controls using fast formula

dist2_between <- function(A, B){
  # A: nA x p, B: nB x p
  # ||a-b||^2 = ||a||^2 + ||b||^2 - 2 a b'
  aa <- rowSums(A^2)
  bb <- rowSums(B^2)
  cross <- A %*% t(B)
  outer(aa, bb, "+") - 2 * cross
}

# Get M nearest indices in each row of a distance matrix

topM_cols <- function(D, M){
  t(apply(D, 1, function(row) order(row, decreasing=FALSE)[1:M]))
}

# Compute ATE  mean(Yhat(1) - Yhat(0))

ate_from_matches <- function(Y, Z, match_ctrl_for_treat, match_treat_for_ctrl){

  nT <- sum(Z==1); nC <- sum(Z==0)

  Yhat1 <- numeric(length(Y))
  Yhat0 <- numeric(length(Y))

  # treated units: observe Y(1), impute Y(0)
  tr_idx <- which(Z==1)
  Yhat1[tr_idx] <- Y[tr_idx]
  Yhat0[tr_idx] <- rowMeans(matrix(Y[match_ctrl_for_treat], nrow=nT))

  # control units: observe Y(0), impute Y(1)
  co_idx <- which(Z==0)
  Yhat0[co_idx] <- Y[co_idx]
  Yhat1[co_idx] <- rowMeans(matrix(Y[match_treat_for_ctrl], nrow=nC))

  mean(Yhat1 - Yhat0)
}

# Max reuse count K_M(i) = number of times unit i is used as a match for others
max_reuse <- function(match_mat, pool_indices){
  # match_mat is n x M, entries are indices into full sample
  tab <- table(factor(as.vector(match_mat), levels = pool_indices))
  max(as.integer(tab))
}

```



## Part 2 Q2a

Distance in prompt;

$$d(i,i^{'}) = \sqrt{\sum_{p=1}^P \dfrac{(X_{ip} - X_{i^{'}p})}{s_p^2}}$$


We implement nearest neighbors from the opposite treatment group for each unit; 

```{r}

# Own covariate matching (normalized Euclidean distance)

own_covariate_matching <- function(Y, Z, X, M){
  sc <- scaled_X_for_covmatch(X)
  Xs <- sc$Xs

  tr_idx <- which(Z==1)
  co_idx <- which(Z==0)
  Xt <- Xs[tr_idx, , drop=FALSE]
  Xc <- Xs[co_idx, , drop=FALSE]

  D2_tc <- dist2_between(Xt, Xc)        

  # For each treated, choose M controls
  nn_ctrl_cols <- topM_cols(D2_tc, M)   
  match_ctrl_for_treat <- matrix(co_idx[nn_ctrl_cols], nrow=nrow(nn_ctrl_cols))

  # For each control, choose M treated 
  D2_ct <- t(D2_tc)                     
  nn_tr_cols <- topM_cols(D2_ct, M)     
  match_treat_for_ctrl <- matrix(tr_idx[nn_tr_cols], nrow=nrow(nn_tr_cols))

  ate_hat <- ate_from_matches(Y, Z, match_ctrl_for_treat, match_treat_for_ctrl)

  # reuse counts across all matchings 
  max_reuse_ctrl <- max_reuse(match_ctrl_for_treat, pool_indices = co_idx)
  max_reuse_tr   <- max_reuse(match_treat_for_ctrl, pool_indices = tr_idx)
  max_reuse_all  <- max(max_reuse_ctrl, max_reuse_tr)

  list(ate = ate_hat, max_reuse = max_reuse_all)
}

Ms <- c(1,4,8)
res_2a <- do.call(rbind, lapply(Ms, function(M){
  out <- own_covariate_matching(Y, Z, Xmm, M)
  data.frame(method="Own covariate match", M=M, ATE_hat=out$ate, max_reuse=out$max_reuse)
}))
print(res_2a)


```

 
 **ATE estimator**
 
 Given matches $M_i$, the prompt's imputation is:
 
  - If $Z_i = 1: \hat{Y}_i(1) = Y_i, \hat{Y}(0) = \frac{1}{M} \sum_{j \in M_i} Y_j$
  
  - If $Z_i = 0: \hat{Y}_i(0) = Y_i, \hat{Y}(1) = \frac{1}{M} \sum_{j \in M_i} Y_j$
  
  
  
  
ATE :

$$\hat{\tau}_M^{ATE} = \frac{1}{N} \sum_{i=1}^N (\hat{Y}_i(1) - \hat{Y}_i(0))$$

Reuse Count:

$$K_M (i) = \sum_{j=1}^N 1 \{i \in M_j\}$$ , $$\text{max}_i K_M(i)$$
## Part 2 Q2b

```{r}

# Own propensity score matching with |e_i - e_j|

own_ps_matching <- function(Y, Z, X, M){
  # full-rank fit for PS
  keep <- qr(X)$pivot[seq_len(qr(X)$rank)]
  Xr <- X[, keep, drop=FALSE]
  fit_ps <- glm.fit(x=Xr, y=Z, family=binomial())
  ehat <- as.numeric(plogis(Xr %*% fit_ps$coefficients))

  tr_idx <- which(Z==1)
  co_idx <- which(Z==0)
  et <- ehat[tr_idx]
  ec <- ehat[co_idx]

  # distance matrices 
  D_tc <- abs(outer(et, ec, "-"))   
  nn_ctrl_cols <- topM_cols(D_tc, M)
  match_ctrl_for_treat <- matrix(co_idx[nn_ctrl_cols], nrow=nrow(nn_ctrl_cols))

  D_ct <- t(D_tc)                   
  nn_tr_cols <- topM_cols(D_ct, M)
  match_treat_for_ctrl <- matrix(tr_idx[nn_tr_cols], nrow=nrow(nn_tr_cols))

  ate_hat <- ate_from_matches(Y, Z, match_ctrl_for_treat, match_treat_for_ctrl)

  max_reuse_ctrl <- max_reuse(match_ctrl_for_treat, pool_indices = co_idx)
  max_reuse_tr   <- max_reuse(match_treat_for_ctrl, pool_indices = tr_idx)
  max_reuse_all  <- max(max_reuse_ctrl, max_reuse_tr)

  list(ate = ate_hat, max_reuse = max_reuse_all, ehat = ehat)
}

res_2b <- do.call(rbind, lapply(Ms, function(M){
  out <- own_ps_matching(Y, Z, Xmm, M)
  data.frame(method="Own PS match", M=M, ATE_hat=out$ate, max_reuse=out$max_reuse)
}))
print(res_2b)

```


## Part 2 Q2c

```{r}

# Using Matching package: covariate + PS matching, M in {1,4,8}

if(!requireNamespace("Matching", quietly=TRUE)) install.packages("Matching")
library(Matching)

# For Matching::Match, X must be numeric matrix.
# Covariate matching: use Xmm variable.
# PS matching: use ehat from PS model above.

# compute ehat 
keep <- qr(Xmm)$pivot[seq_len(qr(Xmm)$rank)]
Xr <- Xmm[, keep, drop=FALSE]
fit_ps <- glm.fit(x=Xr, y=Z, family=binomial())
ehat <- as.numeric(plogis(Xr %*% fit_ps$coefficients))



match_pkg_run <- function(Y, Z, X, M){
  mout <- Match(Y=Y, Tr=Z, X=X, M=M, replace=TRUE, estimand="ATE")

  data.frame(
    M   = M,
    est = as.numeric(mout$est),
    se  = as.numeric(mout$se)
  )
}

res_cov_pkg <- do.call(rbind, lapply(Ms, function(M) match_pkg_run(Y, Z, Xmm, M)))
res_cov_pkg$method <- "Matching pkg: covariate"

res_ps_pkg  <- do.call(rbind, lapply(Ms, function(M) match_pkg_run(Y, Z, matrix(ehat, ncol=1), M)))
res_ps_pkg$method <- "Matching pkg: PS"

res_2c <- rbind(res_cov_pkg, res_ps_pkg)
res_2c <- res_2c[, c("method","M","est","se")]
print(res_2c)


mout <- Match(Y=Y, Tr=Z, X=Xmm, M=1, replace=TRUE, estimand="ATE")
summary(mout)  
mout$est; mout$se
```


```{r}
#
# Compare estimates (own vs package)

cmp <- merge(
  rbind(
    transform(res_2a, est=ATE_hat)[, c("method","M","est","max_reuse")],
    transform(res_2b, est=ATE_hat)[, c("method","M","est","max_reuse")]
  ),
  res_2c,
  by=c("M"),
  all=TRUE,
  suffixes=c(".own",".pkg")
)
print(cmp)


```
**comment**


 - Matching library uses its own variance estimator but my implementation do not compute SEs.
 
 - The Package may break ties differently and uses internal ordering; small differences in matches can shift the estimate.
 
 - My covariate distance uses $SD(diff) = \sqrt{2}*sd(x)$ scaling.
 
 - With replacement, high max_reuse indicates a few good controls/treated are reused often meaning lower bias and higher variance.


## Part 2 Q3

```{r}
df <- read.table("D:/WashU/PHD/Year 2/Sem 2/SDS 5320/Homework/PS2_data.txt", header = TRUE)
df$Z <- as.integer(df$pg == 1)   
df$Y <- df$i_aqoc                

# Pre-treatment covariates
Xvars <- c("i_age","i_sex","i_race","i_educ","i_insu","i_drug","i_seve",
           "com_t","pcs_sd","mcs_sd")

dat <- df[, c("Y","Z",Xvars)]
dat <- dat[complete.cases(dat), ]

# Make characters 
for(v in Xvars){
  if(is.character(dat[[v]])) dat[[v]] <- factor(dat[[v]])
  if(is.factor(dat[[v]])) dat[[v]] <- factor(dat[[v]], levels = levels(dat[[v]]))
}

Y <- dat$Y
Z <- dat$Z
N <- nrow(dat)

# Numeric design matrix for distances
Xmm <- model.matrix(as.formula(paste0("~ ", paste(Xvars, collapse=" + "))), data=dat)
Xmm <- Xmm[, -1, drop=FALSE] 
Xmm <- as.matrix(Xmm); storage.mode(Xmm) <- "double"

Ms <- c(1,4,8)


# distances + nearest neighbors (with replacement)


# Prompt distance uses s_p = SD( X_ip - X_i'p ) across all pairs
# For any variable, SD(diff) = sqrt(2)*sd(x)
scaled_X_for_covmatch <- function(X){
  s <- apply(X, 2, sd)
  s_diff <- sqrt(2) * s
  s_diff[s_diff == 0] <- 1
  Xs <- sweep(X, 2, s_diff, "/")
  list(Xs = Xs, s_diff = s_diff)
}

dist2_between <- function(A, B){
  aa <- rowSums(A^2)
  bb <- rowSums(B^2)
  outer(aa, bb, "+") - 2 * (A %*% t(B))
}

topM_cols <- function(D, M){
  t(apply(D, 1, function(row) order(row, decreasing=FALSE)[1:M]))
}

max_reuse <- function(match_mat, pool_indices){
  tab <- table(factor(as.vector(match_mat), levels = pool_indices))
  max(as.integer(tab))
}


# Build matches 
make_matches_covariate <- function(Z, X, M){
  sc <- scaled_X_for_covmatch(X)
  Xs <- sc$Xs

  tr_idx <- which(Z==1)
  co_idx <- which(Z==0)

  Xt <- Xs[tr_idx, , drop=FALSE]
  Xc <- Xs[co_idx, , drop=FALSE]

  D2_tc <- dist2_between(Xt, Xc)             
  nn_ctrl_cols <- topM_cols(D2_tc, M)
  match_ctrl_for_treat <- matrix(co_idx[nn_ctrl_cols], nrow=nrow(nn_ctrl_cols))

  D2_ct <- t(D2_tc)                          
  nn_tr_cols <- topM_cols(D2_ct, M)
  match_treat_for_ctrl <- matrix(tr_idx[nn_tr_cols], nrow=nrow(nn_tr_cols))

  list(tr_idx=tr_idx, co_idx=co_idx,
       M=M,
       match_ctrl_for_treat=match_ctrl_for_treat,
       match_treat_for_ctrl=match_treat_for_ctrl)
}

make_matches_ps <- function(Z, X, M){
  # PS model: logistic on X (full rank)
  keep <- qr(X)$pivot[seq_len(qr(X)$rank)]
  Xr <- X[, keep, drop=FALSE]
  fit <- glm.fit(x=Xr, y=Z, family=binomial())
  ehat <- as.numeric(plogis(Xr %*% fit$coefficients))

  tr_idx <- which(Z==1)
  co_idx <- which(Z==0)

  et <- ehat[tr_idx]; ec <- ehat[co_idx]
  D_tc <- abs(outer(et, ec, "-"))            
  nn_ctrl_cols <- topM_cols(D_tc, M)
  match_ctrl_for_treat <- matrix(co_idx[nn_ctrl_cols], nrow=nrow(nn_ctrl_cols))

  D_ct <- t(D_tc)                             
  nn_tr_cols <- topM_cols(D_ct, M)
  match_treat_for_ctrl <- matrix(tr_idx[nn_tr_cols], nrow=nrow(nn_tr_cols))

  list(tr_idx=tr_idx, co_idx=co_idx,
M=M, ehat=ehat,

match_ctrl_for_treat=match_ctrl_for_treat,
match_treat_for_ctrl=match_treat_for_ctrl)
}


# Bias-corrected imputation + ATE

# Need mu0_hat(Xi) = E[Y|X,Z=0] and mu1_hat(Xi) = E[Y|X,Z=1]
# Using linear regression 
fit_mu_models <- function(dat, Xvars){
  f <- as.formula(paste0("Y ~ ", paste(Xvars, collapse=" + ")))
  fit0 <- lm(f, data=dat[dat$Z==0, , drop=FALSE])
  fit1 <- lm(f, data=dat[dat$Z==1, , drop=FALSE])
  list(fit0=fit0, fit1=fit1)
}

predict_mu <- function(fit0, fit1, dat){
  mu0 <- as.numeric(predict(fit0, newdata=dat))
  mu1 <- as.numeric(predict(fit1, newdata=dat))
  list(mu0=mu0, mu1=mu1)
}

# Compute bias-corrected ATE per prompt:
# For treated i:  Y~(0) = avg_j [Yj + mu0(Xi) - mu0(Xj)], Y~(1)=Yi
# For control i:  Y~(1) = avg_j [Yj + mu1(Xi) - mu1(Xj)], Y~(0)=Yi
ate_bias_corrected <- function(Y, Z, matches, mu0, mu1){
  tr_idx <- matches$tr_idx
  co_idx <- matches$co_idx
  M <- matches$M

  mC <- matches$match_ctrl_for_treat  
  mT <- matches$match_treat_for_ctrl 

  # Treated units:
  # Ytilde1 = Yi
  # Ytilde0 = avg_j [Yj + mu0_i - mu0_j]
  mu0_i <- mu0[tr_idx]
  mu0_j <- matrix(mu0[mC], nrow=length(tr_idx), ncol=M)
  Y_j0  <- matrix(Y[mC],  nrow=length(tr_idx), ncol=M)
  Ytilde0_tr <- rowMeans(Y_j0 + mu0_i - mu0_j)
  Ytilde1_tr <- Y[tr_idx]

  # Control units:
  # Ytilde0 = Yi
  # Ytilde1 = avg_j [Yj + mu1_i - mu1_j]
  mu1_i <- mu1[co_idx]
  mu1_j <- matrix(mu1[mT], nrow=length(co_idx), ncol=M)
  Y_j1  <- matrix(Y[mT],  nrow=length(co_idx), ncol=M)
  Ytilde1_co <- rowMeans(Y_j1 + mu1_i - mu1_j)
  Ytilde0_co <- Y[co_idx]

  # combined
  Ytilde1 <- numeric(length(Y))
  Ytilde0 <- numeric(length(Y))
  Ytilde1[tr_idx] <- Ytilde1_tr
  Ytilde0[tr_idx] <- Ytilde0_tr
  Ytilde1[co_idx] <- Ytilde1_co
  Ytilde0[co_idx] <- Ytilde0_co

  mean(Ytilde1 - Ytilde0)
}

```


**Bias-Corrected Matching estimator**

We fit $\hat{\mu}_0(X) = E[Y|X,Z = 0]$ and $\hat{\mu}_1(X) = E[Y|X,Z = 1]$ using linear regression. Y is binary





To compute bias-corrected imputed potential outcome and $\hat{\tau}_M^{bcm}$. Given matches $M_i$ from the opposite group then;

 _ If $Z_i = 1:$
 
 $$\hat{Y}_i(1) = Y_i,  \hat{Y}_i(0) = \frac{1}{M} \sum_{j \in M} [Y_j + \hat{\mu}_0(X_i) - \hat{\mu}_0(X_j)]$$
 
 
 
 _ If $Z_i = 0:$
 
 $$\hat{Y}_i(0) = Y_i,  \hat{Y}_i(1) = \frac{1}{M} \sum_{j \in M} [Y_j + \hat{\mu}_1(X_i) - \hat{\mu}_1(X_j)]$$
 
 
then:

$$\hat{\tau}_M^{bcm} = \frac{1}{N} \sum_{i=1}^N (\hat{Y}_i(1) - \hat{Y}_i(0))$$
## Part 3 Q3b

**Bias-corrected as compared to Q2a**

```{r}

# Bias-corrected matching using covariate matches

mu_fits <- fit_mu_models(dat, Xvars)
mu_hat  <- predict_mu(mu_fits$fit0, mu_fits$fit1, dat)
mu0 <- mu_hat$mu0
mu1 <- mu_hat$mu1

res_3a <- lapply(Ms, function(M){
  matches <- make_matches_covariate(Z, Xmm, M)
  tau_bcm <- ate_bias_corrected(Y, Z, matches, mu0, mu1)

  maxR <- max(
max_reuse(matches$match_ctrl_for_treat, matches$co_idx),

max_reuse(matches$match_treat_for_ctrl, matches$tr_idx)
  )

  data.frame(method="Own BCM: covariate match", M=M, tau_bcm=tau_bcm, max_reuse=maxR)
}) |> bind_rows()

knitr::kable(res_3a, digits=4,
             
caption="Bias-corrected matching using covariate matching (own implementation)")
```


Part 2 Q3b

**Bias-correct as compared to Q2b**

```{r}

# Bias-corrected matching using PS matches

res_3b <- lapply(Ms, function(M){
  matches <- make_matches_ps(Z, Xmm, M)
  tau_bcm <- ate_bias_corrected(Y, Z, matches, mu0, mu1)

  maxR <- max(
max_reuse(matches$match_ctrl_for_treat, matches$co_idx),

max_reuse(matches$match_treat_for_ctrl, matches$tr_idx)
  )

data.frame(method="Own BCM: PS match",
M=M, tau_bcm=tau_bcm, max_reuse=maxR)
}) |> bind_rows()

knitr::kable(res_3b, digits=4,

caption="Bias-corrected matching using propensity score matching (own implementation)")

```




## Part 2 Q3c

For M = 4: average absolute bias correction term, we want to compute 

$$\frac{1}{N} \sum_{Z_i = 1} | \hat{\mu}_0(X_i) - \frac{1}{M} \sum_{j \in M_i} \hat{\mu}_0(X_j)|$$

For both covariate and PS matching;

```{r}

# average absolute bias correction term for treated units, M=4:
# (1/N1) sum_{Zi=1} | mu0(Xi) - (1/M) sum_{j in Mi} mu0(Xj) |
avg_abs_bias_term_treated <- function(Z, matches, mu0){
  tr_idx <- matches$tr_idx
  M <- matches$M
  mC <- matches$match_ctrl_for_treat

  mu0_i <- mu0[tr_idx]
  mu0_bar_j <- rowMeans(matrix(mu0[mC], nrow=length(tr_idx), ncol=M))
  mean(abs(mu0_i - mu0_bar_j))
}


# For M=4: average absolute bias correction term (treated)

matches_cov_M4 <- make_matches_covariate(Z, Xmm, M=4)
matches_ps_M4  <- make_matches_ps(Z, Xmm, M=4)

bias_cov <- avg_abs_bias_term_treated(Z, matches_cov_M4, mu0)
bias_ps  <- avg_abs_bias_term_treated(Z, matches_ps_M4,  mu0)

res_3c <- data.frame(
  method = c("Covariate matching","Propensity score matching"),
  M = 4,
  avg_abs_bias_term = c(bias_cov, bias_ps)
)

knitr::kable(res_3c, digits=4,

caption="Average absolute bias correction term for treated units (M=4)")

# Plot for
ggplot(res_3c, aes(x=method, y=avg_abs_bias_term)) +
geom_col() +
labs(x=NULL, y="Average absolute bias correction term",
title="M=4: Bias correction magnitude (treated units)") +
theme_bw() +
theme(axis.text.x = element_text(angle=30, hjust=1))
```

**Comment**

Smaller average absolute term implies matched controls have $\hat{\mu}_0(X)$ closer to treated units. That means better match quality (in terms of the outcome model for contorls), hence smaller bias correction needed. The covariate matching produces smaller bias correction based on the plot.



## 3d

```{r}

# Verify using Matching package with BiasAdjust=TRUE
# PS scores once for package runs
keep <- qr(Xmm)$pivot[seq_len(qr(Xmm)$rank)]
Xr <- Xmm[, keep, drop=FALSE]
fit_ps <- glm.fit(x=Xr, y=Z, family=binomial())
ehat <- as.numeric(plogis(Xr %*% fit_ps$coefficients))

match_pkg_biasadj <- function(Y, Z, X, M){
  mout <- Match(Y=Y, Tr=Z, X=X, M=M, replace=TRUE, estimand="ATE", BiasAdjust=TRUE)
  data.frame(M=M, est=as.numeric(mout$est), se=as.numeric(mout$se))
}

res_pkg_cov_ba <- lapply(Ms, function(M) match_pkg_biasadj(Y, Z, Xmm, M)) |> bind_rows()
res_pkg_cov_ba$method <- "Matching pkg (BiasAdjust=TRUE): covariate"

res_pkg_ps_ba <- lapply(Ms, function(M) match_pkg_biasadj(Y, Z, matrix(ehat, ncol=1), M)) |> bind_rows()
res_pkg_ps_ba$method <- "Matching pkg (BiasAdjust=TRUE): PS"

res_3d <- bind_rows(res_pkg_cov_ba, res_pkg_ps_ba) |>
  dplyr::select(method, M, est, se)

knitr::kable(res_3d, digits=4,
  
caption=" Matching package verification with BiasAdjust=TRUE (estimate and SE)")

```


```{r}

# Comparison own BCM vs package BiasAdjust

own_all <- bind_rows(res_3a, res_3b) |>
  transmute(method, M, est=tau_bcm)

pkg_all <- res_3d |>
  transmute(method, M, est=est, se=se)

cmp <- bind_rows(own_all, pkg_all) |>
  arrange(M, method)

knitr::kable(cmp, digits=4,

caption="Comparison: Own bias-corrected estimates vs Matching::Match(BiasAdjust=TRUE)")

# Plot: estimates vs M for each method family
plot_df <- cmp |>
  mutate(method = factor(method, levels=unique(method)))

ggplot(plot_df, aes(x=factor(M), y=est, group=method)) +
  geom_line() +
  geom_point(size=2) +
  labs(x="M (neighbors)", y="Estimated ATE",
title="Bias-corrected matching estimates across M") +
theme_bw() +
theme(legend.position="bottom")



```


**Comments**

The differences is because;The estimated means are fit once using linear regression on all covariates(full X, with factors). Matching implements the Abadie-Imbens bias correction tied to the covariates supplied in X and its internal procedure. PS matching + Bias_Adjust in matching: If we pass only X = e_hat, its bias adjustment can only use PS not  the full covariates set - so it can differ from the full covariate regression correction. Specifically;

3(c): smaller avg abs bias term => matches are closer in terms of mu0(X), implying higher match quality
3(d): differences between your BCM and Matching::Match(BiasAdjust=TRUE) can come from
  1. tie-breaking / neighbor ordering, 2. how regression adjustment is implemented internally, 3. different handling of rank-deficiency or covariate scaling.


## Part 2 Q4

```{r}

# treatment and outcome
df$Z <- ifelse(df$pg == 1, 1, 0)   
df$Y <- df$i_aqoc

# factors
fac_vars <- c("i_sex","i_race","i_educ","i_insu","i_drug","i_seve")
df[fac_vars] <- lapply(df[fac_vars], factor)


df$i_educ <- as.character(df$i_educ); df$i_educ[df$i_educ=="1"] <- "2"; df$i_educ <- factor(df$i_educ)
df$i_insu <- as.character(df$i_insu); df$i_insu[df$i_insu=="3"] <- "2"; df$i_insu <- factor(df$i_insu)

# numeric covariate matrix
X_cat <- model.matrix(~ 0 + i_sex + i_race + i_educ + i_insu + i_drug + i_seve, data=df)
X <- cbind(i_age=df$i_age, com_t=df$com_t, pcs_sd=df$pcs_sd, mcs_sd=df$mcs_sd, X_cat)

# robust s_p for standardized Euclidean distance 
sp <- sqrt(2 * apply(X, 2, var, na.rm=TRUE))
sp[!is.finite(sp) | sp <= 0] <- 1

X_scaled <- sweep(X, 2, sp, "/")

# propensity score for PS matching
ps_mod <- glm(
  Z ~ i_age + i_sex + i_race + i_educ + i_insu + i_drug + i_seve + com_t + pcs_sd + mcs_sd,
  data=df, family=binomial()
)
df$e_hat <- predict(ps_mod, type="response")
eps <- 1e-6
df$e_hat <- pmin(pmax(df$e_hat, eps), 1-eps)

```



## Part 2 Q4a

**Summary Table**

```{r}
run_match <- function(Y, Z, Xmatch, M, bias_adj) {
  m <- Match(
    Y = Y, Tr = Z, X = Xmatch,
    M = M, replace = TRUE,
    estimand = "ATE",
    BiasAdjust = bias_adj
  )
  tibble(est = m$est, se = m$se.standard)
}

Ms <- c(1,4,8)

res_pkg <- expand_grid(
  match_type = c("Covariate", "PropensityScore"),
  estimator  = c("Simple", "Bias-corrected"),
  M = Ms
) %>%
  rowwise() %>%
  mutate(
    out = list({
      Xuse <- if (match_type=="Covariate") X_scaled else df$e_hat
      run_match(df$Y, df$Z, Xuse, M, bias_adj = (estimator=="Bias-corrected"))
    })
  ) %>%
  ungroup() %>%
  unnest(out) %>%
  mutate(across(c(est,se), ~as.numeric(.)))

res_pkg


tab_summary <- res_pkg %>%
  mutate(method = paste(match_type, estimator, sep=" | ")) %>%
  dplyr::select(method, M, est, se) %>%
  arrange(method, M)

tab_summary

```


```{r}


run_match <- function(Y, Z, Xmatch, M, bias_adj) {
  m <- Match(Y=Y, Tr=Z, X=Xmatch, M=M, replace=TRUE,
             estimand="ATE", BiasAdjust=bias_adj)
  tibble(est=m$est, se=m$se.standard)
}

M_grid <- c(1,2,4,6,8,10,12,15,20)

grid_res <- tidyr::expand_grid(
  match_type = c("Covariate","PropensityScore"),
  estimator  = c("Simple","Bias-corrected"),
  M = M_grid
) %>%
  rowwise() %>%
  mutate(out = list({
    Xuse <- if (match_type=="Covariate") X_scaled else df$e_hat
    run_match(df$Y, df$Z, Xuse, M, bias_adj = (estimator=="Bias-corrected"))
  })) %>%
  ungroup() %>%
  tidyr::unnest(out) %>%
  arrange(match_type, estimator, M)

# Stability diagnostic: range over M-grid
stab <- grid_res %>%
  group_by(match_type, estimator) %>%
  summarise(est_range = max(est) - min(est),
            est_sd = sd(est),
            mean_se = mean(se),
            .groups="drop") %>%
  arrange(est_range)

stab
grid_res


best_method <- stab %>% slice(1) %>% dplyr::select(match_type, estimator)

final_choice <- grid_res %>%
  semi_join(best_method, by=c("match_type","estimator")) %>%
  arrange(se, M) %>%
  slice(1)

final_choice

```


## Part 2 Q4b



```{r}
ggplot(res_pkg, aes(x=M, y=est, group=interaction(match_type, estimator),
                    color=match_type, linetype=estimator)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  labs(x="Number of matches (M)", y="ATE estimate",
       title="ATE vs M for simple and bias-corrected matching",
       color="Matching type", linetype="Estimator")


stability <- res_pkg %>%
  group_by(match_type, estimator) %>%
  summarise(
    est_range = max(est) - min(est),
    est_sd    = sd(est),
    mean_se   = mean(se),
    .groups="drop"
  )
stability

```



## Part 2 Q4c

I think the propensity score with biased corrected matching at M = 4. At this M value, the ATE estimate neither takes too many or too few matching, it makes the estimates between the other also. Based on the plot, choosing the BCM PS means that the value of M here does not affect the overall estimate in a major way as thee estimate relatively.

## Part 2 Q4d

```{r}
M_grid <- c(1,2,4,6,8,10,12,15,20)

grid_res <- expand_grid(
  match_type = c("Covariate", "PropensityScore"),
  estimator  = c("Simple", "Bias-corrected"),
  M = M_grid
) %>%
  rowwise() %>%
  mutate(
    out = list({
      Xuse <- if (match_type=="Covariate") X_scaled else df$e_hat
      run_match(df$Y, df$Z, Xuse, M, bias_adj = (estimator=="Bias-corrected"))
    })
  ) %>%
  ungroup() %>%
  unnest(out)

grid_res

ggplot(grid_res, aes(x=M, y=est, group=interaction(match_type, estimator),
                     color=match_type, linetype=estimator)) +
  geom_point() + geom_line() +
  theme_bw() +
  labs(x="M", y="ATE", title="ATE vs M (extended grid)")

```


```{r}
grid_pick <- grid_res %>%
  group_by(match_type, estimator) %>%
  arrange(M) %>%
  mutate(delta = abs(est - lag(est))) %>%
  summarise(
    M_star = M[which.min(se)],   # or choose based on smallest SE
    est_star = est[which.min(se)],
    se_star  = min(se),
    .groups="drop"
  )

grid_pick

```

**Additional Evidence**

```{r}
logit <- function(p) log(p/(1-p))
sd_logit_ps <- sd(logit(df$e_hat))

cal_grid <- c(0.1, 0.2, 0.3) * sd_logit_ps  # common choices

run_ps_caliper <- function(M, cal) {
  m <- Match(
    Y=df$Y, Tr=df$Z, X=df$e_hat,
    M=M, replace=TRUE, estimand="ATE",
    BiasAdjust=TRUE,
    caliper=cal
  )
  tibble(M=M, caliper=cal, est=m$est, se=m$se.standard)
}

ps_cal_res <- bind_rows(lapply(M_grid, function(M)
  bind_rows(lapply(cal_grid, function(cal) {
    out <- tryCatch(run_ps_caliper(M, cal), error=function(e) NULL)
    out
  }))
))

ps_cal_res

```
