---
title: "Homework 3"
author: "Gifty Osei"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{pdfpages}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, tidy = TRUE,dev = "cairo_pdf")

library(latex2exp)
library(dplyr)
library(knitr)
#library(kableExtra)
library(ggplot2)
library(forcats)
library(tidyr)
library(cobalt)
library(tibble)
library(Matching)
#library(pander)
library(broom)
library(broom.mixed)
```


## Acknowledgement / Disclosure Statement

I discussed this homework with Jacob Antici at some point specifically on some suggested library warnings, however my work is completely independent. I also used Chat-GPT to fine tune my writings and occasionally in helping to understand with errors and fixing bugs in my code.

## Part 1

```{r,include=FALSE}

# Part 1: Double Machine Learning for ATE

library(MASS)

# Build covariance matrix

make_sigma <- function(p = 5, rho = 0.7) {
  Sigma <- matrix(0, nrow = p, ncol = p)
  for (j in 1:p) {
    for (k in 1:p) {
      if (j == k) {
        Sigma[j, k] <- 1
      } else {
        Sigma[j, k] <- rho^abs(j - k)
      }
    }
  }
  Sigma
}


# True nuisance functions

propensity_fun <- function(X) {
  # X is an n x 5 matrix or data frame
  linpred <- X[, 1] + 0.5 * X[, 2]
  exp(linpred) / (1 + exp(linpred))
}

m0_fun <- function(X) {
  X[, 1] + X[, 2]
}

m1_fun <- function(X) {
  X[, 1] + X[, 2] + X[, 1] * X[, 3] + 1
}

# -----------------------------
# True ATE

true_ate <- function() {
  # E[X1] = E[X2] = 0
  # E[X1 X3] = Cov(X1, X3) = 0.7^2 = 0.49
  1 + 0.7^2
}


#  Simulate the dataset

simulate_dml_data <- function(n, seed = 123) {
  set.seed(seed)
  
  p <- 5
  Sigma <- make_sigma(p = p, rho = 0.7)
  
  # Generate X ~ N(0, Sigma)
  X <- MASS::mvrnorm(n = n, mu = rep(0, p), Sigma = Sigma)
  colnames(X) <- paste0("X", 1:p)
  
  # Propensity score
  e <- propensity_fun(X)
  
  # Treatment assignment
  Z <- rbinom(n = n, size = 1, prob = e)
  
  # Potential outcome regressions
  m0 <- m0_fun(X)
  m1 <- m1_fun(X)
  
  # Noise
  eps <- rnorm(n, mean = 0, sd = 1)
  
  # Observed outcome
  Y <- Z * m1 + (1 - Z) * m0 + eps
  
  # Potential outcomes (optional, useful for checking)
  Y0 <- m0 + eps
  Y1 <- m1 + eps
  
  data.frame(
    Y = Y,
    Z = Z,
    X1 = X[, 1],
    X2 = X[, 2],
    X3 = X[, 3],
    X4 = X[, 4],
    X5 = X[, 5],
    e_true = e,
    m0_true = m0,
    m1_true = m1,
    Y0 = Y0,
    Y1 = Y1
  )
}

# test

tau0 <- true_ate()
tau0


dat <- simulate_dml_data(n = 1000, seed = 101)

head(dat)
mean(dat$Z)             # observed treatment rate
mean(dat$Y)
mean(dat$Y1 - dat$Y0)   # samplethe treatment effect
```

```{r, include=FALSE}
# True ATE
true_ate()

# Large sample check
dat_big <- simulate_dml_data(n = 100000, seed = 999)

mean(dat_big$Y1 - dat_big$Y0)

round(cor(dat_big[, c("X1","X2","X3","X4","X5")]), 2)
```



```{r, include=FALSE}
#*Data Simplified*

n <- 1000
p <- 5

Sigma <- outer(1:p, 1:p, function(j, k) 0.7^abs(j - k))
diag(Sigma) <- 1

X <- MASS::mvrnorm(n, mu = rep(0, p), Sigma = Sigma)

e  <- plogis(X[,1] + 0.5 * X[,2])
Z  <- rbinom(n, 1, e)
m0 <- X[,1] + X[,2]
m1 <- X[,1] + X[,2] + X[,1] * X[,3] + 1
Y  <- Z * m1 + (1 - Z) * m0 + rnorm(n)

dat <- data.frame(Y, Z, X1 = X[,1], X2 = X[,2], X3 = X[,3], X4 = X[,4], X5 = X[,5])
```



**Q1**

Now we simulate data using the description given;

```{r}

# Q1 Naive Machine Learning Estimator

library(ranger)
library(ggplot2)
library(dplyr)


# Covariance matrix

make_sigma <- function(p = 5, rho = 0.7) {
  Sigma <- outer(1:p, 1:p, function(j, k) rho^abs(j - k))
  diag(Sigma) <- 1
  Sigma
}


# True nuisance functions

propensity_fun <- function(X) {
  plogis(X[, 1] + 0.5 * X[, 2])
}

m0_fun <- function(X) {
  X[, 1] + X[, 2]
}

m1_fun <- function(X) {
  X[, 1] + X[, 2] + X[, 1] * X[, 3] + 1
}

true_tau <- 1.49
true_mu0 <- 0
true_mu1 <- 1.49


# Simulate dataset

simulate_data <- function(n = 500, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)

  p <- 5
  Sigma <- make_sigma(p = p, rho = 0.7)

  X <- MASS::mvrnorm(n = n, mu = rep(0, p), Sigma = Sigma)
  colnames(X) <- paste0("X", 1:5)

  e <- propensity_fun(X)
  Z <- rbinom(n, size = 1, prob = e)

  m0 <- m0_fun(X)
  m1 <- m1_fun(X)

  eps <- rnorm(n, mean = 0, sd = 1)
  Y <- Z * m1 + (1 - Z) * m0 + eps

  data.frame(
    Y = Y,
    Z = Z,
    X1 = X[, 1],
    X2 = X[, 2],
    X3 = X[, 3],
    X4 = X[, 4],
    X5 = X[, 5],
    m0_true = m0,
    m1_true = m1,
    e_true = e
  )
}
```


```{r}
library(ranger)

# replication

one_rep_naive <- function(n = 500, seed = NULL,
                          num.trees = 500,
                          min.node.size = 5) {

  dat <- simulate_data(n = n, seed = seed)

  # Random split into equal halves
  idx <- sample(seq_len(n))
  I1 <- idx[1:(n / 2)]
  I2 <- idx[(n / 2 + 1):n]

  train <- dat[I1, ]
  est   <- dat[I2, ]

  xvars <- paste0("X", 1:5)
  form <- as.formula(paste("Y ~", paste(xvars, collapse = " + ")))

  # Separate treated/control training sets
  train1 <- train[train$Z == 1, ]
  train0 <- train[train$Z == 0, ]
  if (nrow(train1) < 10 || nrow(train0) < 10) {
    return(NULL)
  }

  # Fit outcome models on traingin set
  fit1 <- ranger(
    formula = form,
    data = train1,
    num.trees = num.trees,
    min.node.size = min.node.size,
    seed = seed
  )

  fit0 <- ranger(
    formula = form,
    data = train0,
    num.trees = num.trees,
    min.node.size = min.node.size,
    seed = seed + 1
  )

  # Predict on test set
  m1_hat_I2 <- predict(fit1, data = est)$predictions
  m0_hat_I2 <- predict(fit0, data = est)$predictions

  # Naive estimators
  mu1_hat <- mean(m1_hat_I2)
  mu0_hat <- mean(m0_hat_I2)
  tau_hat <- mu1_hat - mu0_hat

  # Empirical SE for tau_hat on I2 using plug-in variance
  se_mu1 <- sd(m1_hat_I2) / sqrt(nrow(est))
  se_mu0 <- sd(m0_hat_I2) / sqrt(nrow(est))
  se_tau <- sqrt(se_mu1^2 + se_mu0^2)

  # Quantities for decomposition of sqrt(n)(mu1_hat - mu1)
  # Here P_n is over I2, with n2 = |I2|
  n2 <- nrow(est)

  a_star <- sqrt(n2) * (mean(est$m1_true) - true_mu1)
  b_star <- sqrt(n2) * (mean(m1_hat_I2 - est$m1_true) - mean(m1_hat_I2 - est$m1_true))
  #compute directly as empirical process:
  b_star <- sqrt(n2) * ((mean(m1_hat_I2 - est$m1_true)) -
                          mean(m1_hat_I2 - est$m1_true))

  # The above is identically zero, because the second term is the same quantity.
  # what we want:
  # sqrt(n2) * ( P_n[g] - E[g | training] )
  # We approximate E[g | training] by a large Monte Carlo sample from X
  Sigma <- make_sigma(5, 0.7)
  X_mc <- MASS::mvrnorm(n = 20000, mu = rep(0, 5), Sigma = Sigma)
  colnames(X_mc) <- paste0("X", 1:5)
  X_mc_df <- as.data.frame(X_mc)

  m1_hat_mc <- predict(fit1, data = X_mc_df)$predictions
  m1_true_mc <- m1_fun(as.matrix(X_mc))

  Eg_hat_minus_m1 <- mean(m1_hat_mc - m1_true_mc)

  b_star <- sqrt(n2) * (mean(m1_hat_I2 - est$m1_true) - Eg_hat_minus_m1)
  c_star <- sqrt(n2) * Eg_hat_minus_m1

  data.frame(
    mu1_hat = mu1_hat,
    mu0_hat = mu0_hat,
    tau_hat = tau_hat,
    se_tau = se_tau,
    std_tau = (tau_hat - true_tau) / se_tau,
    a_star = a_star,
    b_star = b_star,
    c_star = c_star
  )
}
```



```{r}

# Run simulation

set.seed(2026)

## bootstrap
B <- 1000
results_list <- vector("list", B)

for (b in 1:B) {
  if (b %% 50 == 0) cat("Replication:", b, "\n")
  results_list[[b]] <- one_rep_naive(n = 500, seed = 1000 + b)
}

results <- bind_rows(results_list)

# remove any NULL/failed reps
results <- results %>% filter(is.finite(std_tau), is.finite(se_tau), se_tau > 0)

summary(results)
```


**1a**

```{r}
results %>%
  summarise(
    mean_mu1_hat = mean(mu1_hat),
    mean_mu0_hat = mean(mu0_hat),
    mean_tau_hat = mean(tau_hat),
    bias_tau_hat = mean(tau_hat - true_tau),
    sd_tau_hat = sd(tau_hat),
    mean_se_tau = mean(se_tau)
  )
```


**1b**

```{r, fig.cap="This figure shows the histogram of the standardized estimates with an overlay of standard normal density."}

ggplot(results, aes(x = std_tau)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 35, color = "black", fill = "grey80") +
  stat_function(fun = dnorm, linewidth = 1) +
  labs(
    title = "Naive ML estimator: standardized tau-hat",
    x = expression((hat(tau)[naive] - tau[0]) / se(hat(tau)[naive])),
    y = "Density"
  ) +
  theme_bw(base_size = 14)
```
**Comment**

We see from the figure that the normal density is shifted to the left of the histogram and not covering but centers at 0.


```{r, fig.cap="This is the QQ-plot of the standardized naive estimator to check the normality assumption."}
mean(results$std_tau)
sd(results$std_tau)
quantile(results$std_tau, probs = c(0.025, 0.5, 0.975))


ggplot(results, aes(sample = std_tau)) +
  stat_qq() +
  stat_qq_line() +
  labs(
    title = "QQ plot of standardized naive estimator"
  ) +
  theme_bw(base_size = 14)
```


**1c**

Now we chech the behavior of $a^*, b^*, c^*$ by summarizing the decomposition terms

```{r}
results %>%
  summarise(
    mean_a = mean(a_star),
    sd_a = sd(a_star),
    mean_b = mean(b_star),
    sd_b = sd(b_star),
    mean_c = mean(c_star),
    sd_c = sd(c_star)
  )
```

```{r}
p1 <- ggplot(results, aes(x = a_star)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, color = "black", fill = "grey80") +
  stat_function(fun = dnorm, linewidth = 1) +
  labs(title = expression(a^"*"), x = expression(a^"*"), y = "Density") +
  theme_bw(base_size = 13)

p2 <- ggplot(results, aes(x = b_star)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, color = "black", fill = "grey80") +
  labs(title = expression(b^"*"), x = expression(b^"*"), y = "Density") +
  theme_bw(base_size = 13)

p3 <- ggplot(results, aes(x = c_star)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, color = "black", fill = "grey80") +
  labs(title = expression(c^"*"), x = expression(c^"*"), y = "Density") +
  theme_bw(base_size = 13)


# p1
# p2
# p3

library(patchwork)
p1 + p2 + p3
```

**Comment**

Based on the results, given that $a^* = \sqrt{n}P_n[m_1(X)-\mu_1]$. This is the average of i.i.d mean-zero terms so the CLT it is approximately normal. The histogram is roughly bell-shaped, centered near 0 and stable variation.

- $b^* = \sqrt{n}(P_n - E)[\hat{m}_1(X)-m_1(X)]$. This is an empirical-process fluctuation term. If the prediction error $\hat{m}_1(X)- m_1(X)$ is sufficiently small in $L_2$, then this term is small and tend to zero. In finite samples, it may not be exactly negligible, but compared with the other terms it is often less problematic.

- Given that $c^* = \sqrt{n}E[\hat{m}_1(X) - m_1(X)]$ is the regularization bias term and the most important. For the naive estimator, unless the ML learner estimates $m_1$ extremely accurately and with sufficiently small bias, this term do not vanish as seen in the results above.


**2**


**a**

```{r, eval=FALSE, include=FALSE}

# Orthogonal Score Without Sample Splitting



# Covariance matrix

make_sigma <- function(p = 5, rho = 0.7) {
  Sigma <- outer(1:p, 1:p, function(j, k) rho^abs(j - k))
  diag(Sigma) <- 1
  Sigma
}

# True nuisance functions

propensity_fun <- function(X) {
  plogis(X[, 1] + 0.5 * X[, 2])
}

m0_fun <- function(X) {
  X[, 1] + X[, 2]
}

m1_fun <- function(X) {
  X[, 1] + X[, 2] + X[, 1] * X[, 3] + 1
}

true_tau <- 1.49
true_mu0 <- 0
true_mu1 <- 1.49


# Simulate dataset again

simulate_data <- function(n = 500, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)

  Sigma <- make_sigma(5, 0.7)
  X <- MASS::mvrnorm(n = n, mu = rep(0, 5), Sigma = Sigma)
  colnames(X) <- paste0("X", 1:5)

  e <- propensity_fun(X)
  Z <- rbinom(n, 1, e)

  m0 <- m0_fun(X)
  m1 <- m1_fun(X)

  eps <- rnorm(n, mean = 0, sd = 1)
  Y <- Z * m1 + (1 - Z) * m0 + eps

  data.frame(
    Y = Y,
    Z = Z,
    X1 = X[,1],
    X2 = X[,2],
    X3 = X[,3],
    X4 = X[,4],
    X5 = X[,5],
    m0_true = m0,
    m1_true = m1,
    e_true = e
  )
}


```

```{r}
one_rep_dr_manual <- function(n = 500, seed = NULL,
                              num.trees = 500,
                              min.node.size = 5,
                              trim = 0.01) {
  dat <- simulate_data(n = n, seed = seed)


  # propensity score model: e(x) = P(Z=1 | X=x)
  dat$Z_factor <- factor(dat$Z, levels = c(0, 1))

  fit_e <- ranger(
    Z_factor ~ X1 + X2 + X3 + X4 + X5,
    data = dat,
    probability = TRUE,
    num.trees = num.trees,
    min.node.size = min.node.size,
    seed = seed
  )

  # m1(x): fit on treated units only
  fit_m1 <- ranger(
    Y ~ X1 + X2 + X3 + X4 + X5,
    data = dat[dat$Z == 1, ],
    num.trees = num.trees,
    min.node.size = min.node.size,
    seed = seed + 1
  )

  # m0(x): fit on control units only
  fit_m0 <- ranger(
    Y ~ X1 + X2 + X3 + X4 + X5,
    data = dat[dat$Z == 0, ],
    num.trees = num.trees,
    min.node.size = min.node.size,
    seed = seed + 2
  )

  e_hat  <- predict(fit_e,  data = dat)$predictions[, "1"]
  m1_hat <- predict(fit_m1, data = dat)$predictions
  m0_hat <- predict(fit_m0, data = dat)$predictions

  # trim estimated propensity scores
  e_hat <- pmin(pmax(e_hat, trim), 1 - trim)

  # individual contributions for mu1_DR
  dr1_i <- dat$Z * (dat$Y - m1_hat) / e_hat + m1_hat

  # individual contributions for mu0_DR
  dr0_i <- (1 - dat$Z) * (dat$Y - m0_hat) / (1 - e_hat) + m0_hat

  # sample averages
  mu1_hat_dr <- mean(dr1_i)
  mu0_hat_dr <- mean(dr0_i)

  # DR ATE estimator
  tau_hat_dr <- mu1_hat_dr - mu0_hat_dr

  # standard error based on estimated influence function
  phi_i <- dr1_i - dr0_i - tau_hat_dr
  se_tau <- sqrt(var(phi_i) / n)

  data.frame(
    mu1_hat_dr = mu1_hat_dr,
    mu0_hat_dr = mu0_hat_dr,
    tau_hat_dr = tau_hat_dr,
    se_tau = se_tau,
    std_tau = (tau_hat_dr - true_tau) / se_tau
  )
}


set.seed(123)
one_rep_dr_manual(n = 500, seed = 101)
```

```{r}
one_rep_dr_manual_full <- function(n = 500, seed = NULL,
                                   num.trees = 500,
                                   min.node.size = 5,
                                   trim = 0.01) {
  dat <- simulate_data(n = n, seed = seed)

  dat$Z_factor <- factor(dat$Z, levels = c(0, 1))

  fit_e <- ranger(
    Z_factor ~ X1 + X2 + X3 + X4 + X5,
    data = dat,
    probability = TRUE,
    num.trees = num.trees,
    min.node.size = min.node.size,
    seed = seed
  )

  fit_m1 <- ranger(
    Y ~ X1 + X2 + X3 + X4 + X5,
    data = dat[dat$Z == 1, ],
    num.trees = num.trees,
    min.node.size = min.node.size,
    seed = seed + 1
  )

  fit_m0 <- ranger(
    Y ~ X1 + X2 + X3 + X4 + X5,
    data = dat[dat$Z == 0, ],
    num.trees = num.trees,
    min.node.size = min.node.size,
    seed = seed + 2
  )

  e_hat  <- predict(fit_e,  data = dat)$predictions[, "1"]
  m1_hat <- predict(fit_m1, data = dat)$predictions
  m0_hat <- predict(fit_m0, data = dat)$predictions

  e_hat <- pmin(pmax(e_hat, trim), 1 - trim)

  dr1_i <- dat$Z * (dat$Y - m1_hat) / e_hat + m1_hat
  dr0_i <- (1 - dat$Z) * (dat$Y - m0_hat) / (1 - e_hat) + m0_hat

  mu1_hat_dr <- mean(dr1_i)
  mu0_hat_dr <- mean(dr0_i)
  tau_hat_dr <- mu1_hat_dr - mu0_hat_dr

  phi_i <- dr1_i - dr0_i - tau_hat_dr
  se_tau <- sqrt(var(phi_i) / n)

  list(
    estimates = data.frame(
      mu1_hat_dr = mu1_hat_dr,
      mu0_hat_dr = mu0_hat_dr,
      tau_hat_dr = tau_hat_dr,
      se_tau = se_tau,
      std_tau = (tau_hat_dr - true_tau) / se_tau
    ),
    dr1_i = dr1_i,
    dr0_i = dr0_i,
    e_hat = e_hat,
    m1_hat = m1_hat,
    m0_hat = m0_hat,
    dat = dat
  )
}


out <- one_rep_dr_manual_full(n = 500, seed = 101)

out$estimates
head(out$dr1_i)
head(out$dr0_i)
head(out$e_hat)
head(out$m1_hat)
head(out$m0_hat)
```

```{r}
set.seed(2026)

B <- 1000
res_q2a <- bind_rows(lapply(1:B, function(b) {
  one_rep_dr_manual(n = 500, seed = 2000 + b)
}))

summary(res_q2a)

res_q2a %>%
  summarise(
    mean_mu1 = mean(mu1_hat_dr),
    mean_mu0 = mean(mu0_hat_dr),
    mean_tau = mean(tau_hat_dr),
    bias_tau = mean(tau_hat_dr - true_tau),
    sd_tau = sd(tau_hat_dr),
    mean_se = mean(se_tau),
    mean_std = mean(std_tau),
    sd_std = sd(std_tau)
  )

ggplot(res_q2a, aes(x = std_tau)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 35, fill = "grey80", color = "black") +
  stat_function(fun = dnorm, linewidth = 1) +
  labs(
    title = "Q2(a): DR estimator without sample splitting",
    x = expression((hat(tau)[DR] - tau[0]) / se(hat(tau)[DR])),
    y = "Density"
  ) +
  theme_bw(base_size = 14)
```

**b**

```{r}

# One replication for manual DR

one_rep_dr_manual <- function(n = 500, seed = NULL,
                              num.trees = 500,
                              min.node.size = 5,
                              trim = 0.01) {

  dat <- simulate_data(n = n, seed = seed)

  # propensity model on full sample
  dat$Z_factor <- factor(dat$Z, levels = c(0, 1))

  fit_e <- ranger(
    Z_factor ~ X1 + X2 + X3 + X4 + X5,
    data = dat,
    num.trees = num.trees,
    min.node.size = min.node.size,
    probability = TRUE,
    seed = seed
  )

  # outcome models on full sample
  dat1 <- dat[dat$Z == 1, ]
  dat0 <- dat[dat$Z == 0, ]

  fit_m1 <- ranger(
    Y ~ X1 + X2 + X3 + X4 + X5,
    data = dat1,
    num.trees = num.trees,
    min.node.size = min.node.size,
    seed = seed + 1
  )

  fit_m0 <- ranger(
    Y ~ X1 + X2 + X3 + X4 + X5,
    data = dat0,
    num.trees = num.trees,
    min.node.size = min.node.size,
    seed = seed + 2
  )

  # predictions on full sample
  e_hat  <- predict(fit_e,  data = dat)$predictions[, "1"]
  m1_hat <- predict(fit_m1, data = dat)$predictions
  m0_hat <- predict(fit_m0, data = dat)$predictions

  # trim propensity scores
  e_hat <- pmin(pmax(e_hat, trim), 1 - trim)

  # DR scores for mu1 and mu0
  psi1_i <- dat$Z * (dat$Y - m1_hat) / e_hat + m1_hat
  psi0_i <- (1 - dat$Z) * (dat$Y - m0_hat) / (1 - e_hat) + m0_hat

  mu1_hat <- mean(psi1_i)
  mu0_hat <- mean(psi0_i)
  tau_hat <- mu1_hat - mu0_hat

  # influence-function-based SE
  phi_i <- psi1_i - psi0_i - tau_hat
  se_tau <- sqrt(var(phi_i) / n)

  data.frame(
    mu1_hat = mu1_hat,
    mu0_hat = mu0_hat,
    tau_hat = tau_hat,
    se_tau = se_tau,
    std_tau = (tau_hat - true_tau) / se_tau
  )
}
```


The orthogonal-score estimator is an averageof estimated pseudo-outcomes. So a natural asymptotic variance estimator is based on the empirical variance of

$$\phi_i  = [\frac{Z_i(Y_i - \hat{m}_1(X_i))}{\hat{e}(X_i)} + \hat{m}_1(X_i) ] - [\frac{(1-Z_i)(Y_i - \hat{m}_0(X_i))}{1 - \hat{e}(X_i)} + \hat{m}_0(X_i)] - \hat{\tau}_{DR}$$
Then $\hat{se} (\hat{\tau}_{DR}) = \sqrt{\frac{\widehat{Var}(\phi_i)}{n}}$

```{r}

# Monte Carlo for manual DR

set.seed(2026)

B <- 1000
res_manual_list <- vector("list", B)

for (b in 1:B) {
  if (b %% 50 == 0) cat("Manual DR replication:", b, "\n")
  res_manual_list[[b]] <- one_rep_dr_manual(n = 500, seed = 2000 + b)
}

res_manual <- bind_rows(res_manual_list) %>%
  filter(is.finite(std_tau), is.finite(se_tau), se_tau > 0)

summary(res_manual)
```



```{r}
ggplot(res_manual, aes(x = std_tau)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 35, color = "black", fill = "grey80") +
  stat_function(fun = dnorm, linewidth = 1) +
  labs(
    title = "Orthogonal score without sample splitting (manual)",
    x = expression((hat(tau)[DR] - tau[0]) / se(hat(tau)[DR])),
    y = "Density"
  ) +
  theme_bw(base_size = 14)
```


```{r}
res_manual %>%
  summarise(
    mean_tau_hat = mean(tau_hat),
    bias_tau_hat = mean(tau_hat - true_tau),
    sd_tau_hat = sd(tau_hat),
    mean_se_tau = mean(se_tau),
    mean_std_tau = mean(std_tau),
    sd_std_tau = sd(std_tau)
  )
```



**Verifying with doubleML package**


```{r}
library(DoubleML)
library(mlr3)
library(mlr3learners)
library(data.table)
```


```{r}

# One replication for DoubleML verification

one_rep_dr_doubleml <- function(n = 500, seed = NULL) {

  dat <- simulate_data(n = n, seed = seed)

  dml_data <- DoubleMLData$new(
    data = data.table(dat[, c("Y", "Z", "X1", "X2", "X3", "X4", "X5")]),
    y_col = "Y",
    d_cols = "Z"
  )

  # learners
  ml_g <- lrn("regr.ranger",
              num.trees = 500,
              min.node.size = 5)

  ml_m <- lrn("classif.ranger",
              num.trees = 500,
              min.node.size = 5,
              predict_type = "prob")

  # ATE estimator
  dml_plr <- DoubleMLIRM$new(
    data = dml_data,
    ml_g = ml_g,
    ml_m = ml_m,
    n_folds = 2,
    n_rep = 1,
    score = "ATE"
  )

  # disable cross-fitting / sample splitting as much as package allows
  dml_plr$fit(store_predictions = TRUE)

  tau_hat <- as.numeric(dml_plr$coef)
  se_tau  <- as.numeric(dml_plr$se)

  data.frame(
    tau_hat = tau_hat,
    se_tau = se_tau,
    std_tau = (tau_hat - true_tau) / se_tau
  )
}
```

```{r}

# Monte Carlo DoubleML

set.seed(2027)

B <- 1000
res_dml_list <- vector("list", B)

for (b in 1:B) {
  if (b %% 50 == 0) cat("DoubleML replication:", b, "\n")
  tmp <- try(one_rep_dr_doubleml(n = 500, seed = 3000 + b), silent = TRUE)
  if (!inherits(tmp, "try-error")) {
    res_dml_list[[b]] <- tmp
  }
}

res_dml <- bind_rows(res_dml_list) %>%
  filter(is.finite(std_tau), is.finite(se_tau), se_tau > 0)

summary(res_dml)
```


**Both plots**

```{r}
p_manual <- ggplot(res_manual, aes(x = std_tau)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 35, color = "black", fill = "grey80") +
  stat_function(fun = dnorm, linewidth = 1) +
  labs(
    title = "Manual orthogonal score",
    x = expression((hat(tau)[DR] - tau[0]) / se(hat(tau)[DR])),
    y = "Density"
  ) +
  theme_bw(base_size = 14)

p_dml <- ggplot(res_dml, aes(x = std_tau)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 35, color = "black", fill = "grey80") +
  stat_function(fun = dnorm, linewidth = 1) +
  labs(
    title = "DoubleML verification",
    x = expression((hat(tau)[DR] - tau[0]) / se(hat(tau)[DR])),
    y = "Density"
  ) +
  theme_bw(base_size = 14)


## showing histogram side by side

p_manual + p_dml

```

**Compare both approaches**

```{r}
manual_summary <- res_manual %>%
  summarise(
    mean_tau = mean(tau_hat),
    bias_tau = mean(tau_hat - true_tau),
    sd_tau = sd(tau_hat),
    mean_se = mean(se_tau),
    mean_std = mean(std_tau),
    sd_std = sd(std_tau)
  )

dml_summary <- res_dml %>%
  summarise(
    mean_tau = mean(tau_hat),
    bias_tau = mean(tau_hat - true_tau),
    sd_tau = sd(tau_hat),
    mean_se = mean(se_tau),
    mean_std = mean(std_tau),
    sd_std = sd(std_tau)
  )

manual_summary
dml_summary
```


**Comment on the role of Neyman's orthogonality**

The DR score is orthogonal because the moment condition is constructed so that small errors in estimating the nuisance functions have only second-order effect on the target parameter. meaning, the estimator is much less sensitive to nuisance estimation bias than the naive plug-in estimator. The standardized estimator looks closer to a normal than in Q1 and without splitting the orthogonality helps substantially.

 - 1 naive estimator: no orthogonal correction, sensitive to regularization bias
 
 - 2 Orthogonal score: include bias-correction terms, much more stable and closer to asymptotic normality with small bias estimations.



**3**

**a**

```{r}

# replication for manual DML with 2-fold cross-fitting

one_rep_dml_cf <- function(n = 500, seed = NULL,
                           num.trees = 500,
                           min.node.size = 5,
                           trim = 0.01) {

  dat <- simulate_data(n = n, seed = seed)

  # random 2-fold split
  idx <- sample(seq_len(n))
  fold_id <- rep(1:2, each = n / 2)
  fold_id <- fold_id[order(order(idx))]  # random assignment to folds

  # storage for out-of-fold predictions
  e_hat  <- rep(NA_real_, n)
  m1_hat <- rep(NA_real_, n)
  m0_hat <- rep(NA_real_, n)

  for (k in 1:2) {
    test_idx  <- which(fold_id == k)
    train_idx <- which(fold_id != k)

    train_dat <- dat[train_idx, ]
    test_dat  <- dat[test_idx, ]

    train_dat$Z_factor <- factor(train_dat$Z, levels = c(0, 1))

    # propensity model on training fold
    fit_e <- ranger(
      Z_factor ~ X1 + X2 + X3 + X4 + X5,
      data = train_dat,
      probability = TRUE,
      num.trees = num.trees,
      min.node.size = min.node.size,
      seed = seed + k
    )

    # outcome models on training fold, separately by treatment status
    train1 <- train_dat[train_dat$Z == 1, ]
    train0 <- train_dat[train_dat$Z == 0, ]

    # safety check
    if (nrow(train1) < 10 || nrow(train0) < 10) {
      return(NULL)
    }

    fit_m1 <- ranger(
      Y ~ X1 + X2 + X3 + X4 + X5,
      data = train1,
      num.trees = num.trees,
      min.node.size = min.node.size,
      seed = seed + 100 + k
    )

    fit_m0 <- ranger(
      Y ~ X1 + X2 + X3 + X4 + X5,
      data = train0,
      num.trees = num.trees,
      min.node.size = min.node.size,
      seed = seed + 200 + k
    )

    # out-of-fold predictions
    e_hat[test_idx]  <- predict(fit_e,  data = test_dat)$predictions[, "1"]
    m1_hat[test_idx] <- predict(fit_m1, data = test_dat)$predictions
    m0_hat[test_idx] <- predict(fit_m0, data = test_dat)$predictions
  }

  # trim propensity scores
  e_hat <- pmin(pmax(e_hat, trim), 1 - trim)

  # orthogonal-score pseudo-outcomes
  psi1_i <- dat$Z * (dat$Y - m1_hat) / e_hat + m1_hat
  psi0_i <- (1 - dat$Z) * (dat$Y - m0_hat) / (1 - e_hat) + m0_hat

  mu1_hat <- mean(psi1_i)
  mu0_hat <- mean(psi0_i)
  tau_hat <- mu1_hat - mu0_hat

  # IF-based standard error
  phi_i <- psi1_i - psi0_i - tau_hat
  se_tau <- sqrt(var(phi_i) / n)

  # bonus: CI for mu1
  phi1_i <- psi1_i - mu1_hat
  se_mu1 <- sqrt(var(phi1_i) / n)
  ci_mu1_l <- mu1_hat - 1.96 * se_mu1
  ci_mu1_u <- mu1_hat + 1.96 * se_mu1

  data.frame(
    mu1_hat = mu1_hat,
    mu0_hat = mu0_hat,
    tau_hat = tau_hat,
    se_tau = se_tau,
    std_tau = (tau_hat - true_tau) / se_tau,
    se_mu1 = se_mu1,
    ci_mu1_l = ci_mu1_l,
    ci_mu1_u = ci_mu1_u
  )
}
```

```{r}

# Monte Carlo for manual cross-fitted DML

set.seed(2028)

B <- 1000
res_cf_list <- vector("list", B)

for (b in 1:B) {
  if (b %% 50 == 0) cat("Cross-fitted DML replication:", b, "\n")
  res_cf_list[[b]] <- one_rep_dml_cf(n = 500, seed = 4000 + b)
}

res_cf <- bind_rows(res_cf_list) %>%
  filter(is.finite(std_tau), is.finite(se_tau), se_tau > 0)

summary(res_cf)
```


```{r}
##histogram
p_cf <- ggplot(res_cf, aes(x = std_tau)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 35, fill = "grey80", color = "black") +
  stat_function(fun = dnorm, linewidth = 1) +
  labs(
    title = "Cross-fitted DML estimator (manual)",
    x = expression((hat(tau)[DML] - tau[0]) / se(hat(tau)[DML])),
    y = "Density"
  ) +
  theme_bw(base_size = 14)

p_cf

## summary
res_cf %>%
  summarise(
    mean_tau = mean(tau_hat),
    bias_tau = mean(tau_hat - true_tau),
    sd_tau = sd(tau_hat),
    mean_se = mean(se_tau),
    mean_std = mean(std_tau),
    sd_std = sd(std_tau)
  )
```


**b**

```{r}
## verifying with Dml


# replication DoubleML verification

one_rep_dml_doubleml <- function(n = 500, seed = NULL) {

  dat <- simulate_data(n = n, seed = seed)

  dml_data <- DoubleMLData$new(
    data = data.table(dat[, c("Y", "Z", "X1", "X2", "X3", "X4", "X5")]),
    y_col = "Y",
    d_cols = "Z"
  )

  # learners
  ml_g <- lrn("regr.ranger",
              num.trees = 500,
              min.node.size = 5)

  ml_m <- lrn("classif.ranger",
              num.trees = 500,
              min.node.size = 5,
              predict_type = "prob")

  dml_irm <- DoubleMLIRM$new(
    data = dml_data,
    ml_g = ml_g,
    ml_m = ml_m,
    n_folds = 2,
    n_rep = 1,
    score = "ATE"
  )

  dml_irm$fit()

  tau_hat <- as.numeric(dml_irm$coef)
  se_tau  <- as.numeric(dml_irm$se)
  data.frame(
    tau_hat = tau_hat,
    se_tau = se_tau,
    std_tau = (tau_hat - true_tau) / se_tau
  )
}
```


```{r}
one_rep_dml_cf <- function(n = 500, seed = NULL,
                           num.trees = 500,
                           min.node.size = 5,
                           trim = 0.01) {

  dat <- simulate_data(n = n, seed = seed)

  idx <- sample(seq_len(n))
  folds <- rep(1:2, each = n / 2)
  fold_id <- integer(n)
  fold_id[idx] <- folds

  e_hat  <- rep(NA_real_, n)
  m1_hat <- rep(NA_real_, n)
  m0_hat <- rep(NA_real_, n)

  for (k in 1:2) {
    test_idx  <- which(fold_id == k)
    train_idx <- which(fold_id != k)

    train_dat <- dat[train_idx, ]
    test_dat  <- dat[test_idx, ]

    train_dat$Z_factor <- factor(train_dat$Z, levels = c(0, 1))

    fit_e <- ranger(
      Z_factor ~ X1 + X2 + X3 + X4 + X5,
      data = train_dat,
      probability = TRUE,
      num.trees = num.trees,
      min.node.size = min.node.size,
      seed = seed + k
    )

    fit_m1 <- ranger(
      Y ~ X1 + X2 + X3 + X4 + X5,
      data = train_dat[train_dat$Z == 1, ],
      num.trees = num.trees,
      min.node.size = min.node.size,
      seed = seed + 100 + k
    )

    fit_m0 <- ranger(
      Y ~ X1 + X2 + X3 + X4 + X5,
      data = train_dat[train_dat$Z == 0, ],
      num.trees = num.trees,
      min.node.size = min.node.size,
      seed = seed + 200 + k
    )

    e_hat[test_idx]  <- predict(fit_e, data = test_dat)$predictions[, "1"]
    m1_hat[test_idx] <- predict(fit_m1, data = test_dat)$predictions
    m0_hat[test_idx] <- predict(fit_m0, data = test_dat)$predictions
  }

  e_hat <- pmin(pmax(e_hat, trim), 1 - trim)

  psi1_i <- dat$Z * (dat$Y - m1_hat) / e_hat + m1_hat
  psi0_i <- (1 - dat$Z) * (dat$Y - m0_hat) / (1 - e_hat) + m0_hat

  mu1_hat <- mean(psi1_i)
  mu0_hat <- mean(psi0_i)
  tau_hat <- mu1_hat - mu0_hat

  phi_i <- psi1_i - psi0_i - tau_hat
  se_tau <- sqrt(var(phi_i) / n)

  phi1_i <- psi1_i - mu1_hat
  se_mu1 <- sqrt(var(phi1_i) / n)

  list(
    summary = data.frame(
      mu1_hat = mu1_hat,
      mu0_hat = mu0_hat,
      tau_hat = tau_hat,
      se_tau = se_tau,
      std_tau = (tau_hat - 1.49) / se_tau,
      se_mu1 = se_mu1
    ),
    psi1_i = psi1_i,
    psi0_i = psi0_i,
    e_hat = e_hat,
    m1_hat = m1_hat,
    m0_hat = m0_hat,
    dat = dat
  )
}

out1 <- one_rep_dml_cf(n = 500, seed = 4001)
names(out1)
out1$psi1_i[1:10]
out1$summary
```



```{r}

# Monte Carlo for DoubleML verification

set.seed(2029)

B <- 1000
res_dmlpkg_list <- vector("list", B)

for (b in 1:B) {
  if (b %% 50 == 0) cat("DoubleML package replication:", b, "\n")
  tmp <- try(one_rep_dml_doubleml(n = 500, seed = 5000 + b), silent = TRUE)
  if (!inherits(tmp, "try-error")) {
    res_dmlpkg_list[[b]] <- tmp
  }
}

res_dmlpkg <- bind_rows(res_dmlpkg_list) %>%
  filter(is.finite(std_tau), is.finite(se_tau), se_tau > 0)

summary(res_dmlpkg)
```



```{r}
p_manual_cf <- ggplot(res_cf, aes(x = std_tau)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 35, fill = "grey80", color = "black") +
  stat_function(fun = dnorm, linewidth = 1) +
  labs(
    title = "Manual cross-fitted DML",
    x = expression((hat(tau)[DML] - tau[0]) / se(hat(tau)[DML])),
    y = "Density"
  ) +
  theme_bw(base_size = 14)

p_pkg_cf <- ggplot(res_dmlpkg, aes(x = std_tau)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 35, fill = "grey80", color = "black") +
  stat_function(fun = dnorm, linewidth = 1) +
  labs(
    title = "DoubleML package",
    x = expression((hat(tau)[DML] - tau[0]) / se(hat(tau)[DML])),
    y = "Density"
  ) +
  theme_bw(base_size = 14)

p_manual_cf + p_pkg_cf
```



```{r}
bind_rows(
  results %>%
    summarise(method = "Naive ML",
              mean_tau = mean(tau_hat),
              bias_tau = mean(tau_hat - true_tau),
              sd_tau = sd(tau_hat),
              mean_se = mean(se_tau),
              mean_std = mean(std_tau),
              sd_std = sd(std_tau)),
  res_manual %>%
    summarise(method = "DR no splitting",
              mean_tau = mean(tau_hat),
              bias_tau = mean(tau_hat - true_tau),
              sd_tau = sd(tau_hat),
              mean_se = mean(se_tau),
              mean_std = mean(std_tau),
              sd_std = sd(std_tau)),
  res_cf %>%
    summarise(method = "DML cross-fitted",
              mean_tau = mean(tau_hat),
              bias_tau = mean(tau_hat - true_tau),
              sd_tau = sd(tau_hat),
              mean_se = mean(se_tau),
              mean_std = mean(std_tau),
              sd_std = sd(std_tau))
)
```


**c(Bonus Question)**

\includepdf[pages=1]{D:/WashU/PHD/Year 2/Sem 2/SDS 5320/Homework/Hw3_causal.pdf}

```{r}
## Confidence interval

phi1_i <- out$psi1_i - out$mu1_hat
se_mu1 <- sqrt(var(phi1_i) / n)


## coveragee
# true mu_1 = 1.49
res_cf %>%
  summarise(
    coverage_mu1 = mean(ci_mu1_l <= true_mu1 & ci_mu1_u >= true_mu1),
    avg_len_mu1 = mean(ci_mu1_u - ci_mu1_l)
  )
```






## Part 2

```{r}
library(nnet)
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(stringr)
library(forcats)
```


```{r}
# Load rdata
obj_names <- load("mepsPS3.rdata")
obj_names

# Use the first object if needed
dat <- get(obj_names[1])


# dat <- meps.dat

#str(dat)
#dim(dat)
names(dat)
```


```{r}
# Variables requested
y_var <- "healthExp"
trt_var <- "race"

cont_vars <- c("pcs", "mcs", "age", "bmi", "sinceCheckUp")

cat_vars <- c(
  "male", "married", "health", "mhealth", "anylim", "socialLim", "cogLim",
  "exercise", "hibp", "chd", "emphysema", "cholesterol", "cancer", "stroke",
  "diabetes", "angina", "arthritis", "asthma", "mi", "smoke"
)

all_needed <- c(y_var, trt_var, cont_vars, cat_vars)
setdiff(all_needed, names(dat))
```


**data cleaning**

```{r}
meps <- dat %>%
  dplyr::select(all_of(all_needed)) %>%
  filter(complete.cases(.)) %>%
  mutate(
    race = factor(race),
    race = fct_relevel(race, "White", "Asian", "Black", "Hispanic")
  )

# Convert categorical confounders to factors
meps <- meps %>%
  mutate(across(all_of(cat_vars), ~ as.factor(.)))

# Keep continuous ones numeric
meps <- meps %>%
  mutate(across(all_of(cont_vars), ~ as.numeric(.)),
         healthExp = as.numeric(healthExp))

table(meps$race)
summary(meps$healthExp)
```


**a** 

```{r}
# Multinomial logit formula
gps_formula <- as.formula(
  paste(trt_var, "~", paste(c(cont_vars, cat_vars), collapse = " + "))
)

# Fit GPS model
gps_fit <- nnet::multinom(gps_formula, data = meps, trace = FALSE)

summary(gps_fit)
```


```{r}
gps_hat <- predict(gps_fit, type = "probs")

# Ensure matrix format even if one column edge case happens
gps_hat <- as.matrix(gps_hat)

colnames(gps_hat)
head(gps_hat)

# Attach fitted probabilities
for (lev in levels(meps$race)) {
  meps[[paste0("gps_", lev)]] <- gps_hat[, lev]
}

## observed-treatment Prob for each unit:

meps$gps_obs <- gps_hat[cbind(seq_len(nrow(meps)), match(meps$race, colnames(gps_hat)))]

summary(meps$gps_obs)
```

**checking overlap**

```{r}
meps %>%
  group_by(race) %>%
  summarise(
    n = n(),
    min_gps = min(gps_obs),
    p01 = quantile(gps_obs, 0.01),
    p05 = quantile(gps_obs, 0.05),
    p10 = quantile(gps_obs, 0.10),
    median = median(gps_obs),
    p90 = quantile(gps_obs, 0.90),
    p95 = quantile(gps_obs, 0.95),
    p99 = quantile(gps_obs, 0.99),
    max_gps = max(gps_obs)
  )
```


**plots**

```{r}
ggplot(meps, aes(x = gps_obs, color = race, fill = race)) +
  geom_density(alpha = 0.15) +
  labs(
    title = "Overlap check: distribution of observed-group generalized propensity score",
    x = "Estimated GPS for observed treatment",
    y = "Density"
  ) +
  theme_bw(base_size = 14)



ggplot(meps, aes(x = race, y = gps_obs, fill = race)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Observed-group GPS by race",
    x = "Race",
    y = "Estimated GPS"
  ) +
  theme_bw(base_size = 14) +
  theme(legend.position = "none")
```
**Comment**

We see a somewhat okay overlap between all groups in general but it is weak between some selected 2 groups like Asians vs Black or White vs Asian. 

**b**

Constructing Weights for IPW and overlap;

```{r}
meps$w_ipw <- 1 / meps$gps_obs

summary(meps$w_ipw)
quantile(meps$w_ipw, probs = c(.5, .9, .95, .99))

```

```{r}
gps_mat <- gps_hat
h_x <- 1 / rowSums(1 / gps_mat)

meps$w_ow <- h_x / meps$gps_obs

summary(meps$w_ow)
quantile(meps$w_ow, probs = c(.5, .9, .95, .99))

```


```{r}
## weight helper functions
w_mean <- function(x, w) {
  sum(w * x) / sum(w)
}

w_var <- function(x, w) {
  mu <- w_mean(x, w)
  sum(w * (x - mu)^2) / sum(w)
}
```


```{r}
# Build model matrix of confounders only
X_df <- meps %>%
  dplyr::select(all_of(c(cont_vars, cat_vars)))

X_mm <- model.matrix(~ . - 1, data = X_df)
X_bal <- as.data.frame(X_mm)

colnames(X_bal) <- make.names(colnames(X_bal))
```

**Computing ASD**

```{r}
pairwise_asd <- function(X, A, w = NULL) {
  if (is.null(w)) w <- rep(1, length(A))
  
  groups <- levels(A)
  pairs <- combn(groups, 2, simplify = FALSE)
  
  out <- map_dfr(colnames(X), function(v) {
    x <- X[[v]]
    
    map_dfr(pairs, function(pr) {
      g1 <- pr[1]
      g2 <- pr[2]
      
      i1 <- A == g1
      i2 <- A == g2
      
      m1 <- w_mean(x[i1], w[i1])
      m2 <- w_mean(x[i2], w[i2])
      s1 <- w_var(x[i1], w[i1])
      s2 <- w_var(x[i2], w[i2])
      
      sd_pool <- sqrt((s1 + s2) / 2)
      asd <- ifelse(sd_pool > 0, abs(m1 - m2) / sd_pool, NA_real_)
      
      tibble(
        variable = v,
        group1 = g1,
        group2 = g2,
        asd = asd
      )
    })
  })
  
  out
}
```


**Compute balance tables**

```{r}
bal_unw <- pairwise_asd(X_bal, meps$race, w = rep(1, nrow(meps))) %>%
  mutate(method = "Unweighted")

bal_ipw <- pairwise_asd(X_bal, meps$race, w = meps$w_ipw) %>%
  mutate(method = "IPW")

bal_ow <- pairwise_asd(X_bal, meps$race, w = meps$w_ow) %>%
  mutate(method = "Overlap")

bal_all <- bind_rows(bal_unw, bal_ipw, bal_ow)
```


**Summary**

```{r}
bal_all %>%
  group_by(method) %>%
  summarise(
    max_asd = max(asd, na.rm = TRUE),
    mean_asd = mean(asd, na.rm = TRUE),
    median_asd = median(asd, na.rm = TRUE),
    prop_below_0_1 = mean(asd < 0.1, na.rm = TRUE)
  )

## worst covaraite

bal_all %>%
  group_by(method) %>%
  slice_max(order_by = asd, n = 15) %>%
  arrange(method, desc(asd))



```

```{r}
## plot of worst covariates
ggplot(bal_all, aes(x = asd, fill = method)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  facet_wrap(~ method, scales = "free_y") +
  geom_vline(xintercept = 0.1, linetype = 2) +
  labs(
    title = "Pairwise standardized differences across covariates",
    x = "Absolute standardized difference",
    y = "Count"
  ) + theme(axis.text.x = element_text(angle = 60))+
  theme_bw(base_size = 14)






```


```{r}

## love plot
library(cobalt)


love_df <- bal_all %>%
  group_by(method, variable) %>%
  summarise(max_asd = max(asd, na.rm = TRUE), .groups = "drop") %>%
  group_by(variable) %>%
  mutate(order_var = max(max_asd, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(order_var) %>%
  mutate(variable = factor(variable, levels = unique(variable)))

ggplot(love_df, aes(x = max_asd, y = variable, color = method)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0.1, linetype = 2, color = "red") +
  labs(
    title = "Love Plot: Covariate Balance",
    x = "Maximum Pairwise Standardized Difference",
    y = "Covariate",
    color = "Method"
  ) +
  theme_bw(base_size = 14)
```



```{r}
## effective sample size
ess_fun <- function(w) (sum(w)^2) / sum(w^2)

meps %>%
  group_by(race) %>%
  summarise(
    n = n(),
    ess_ipw = ess_fun(w_ipw),
    ess_ow  = ess_fun(w_ow)
  )
```



```{r}
weighted_group_means <- function(data, y, trt, w) {
  yv <- data[[y]]
  av <- data[[trt]]
  wv <- data[[w]]
  
  tibble(
    race = levels(av),
    mean = map_dbl(levels(av), function(g) {
      idx <- av == g
      sum(wv[idx] * yv[idx]) / sum(wv[idx])
    })
  )
}


mu_ipw <- weighted_group_means(meps, y = "healthExp", trt = "race", w = "w_ipw")
mu_ow  <- weighted_group_means(meps, y = "healthExp", trt = "race", w = "w_ow")

mu_ipw
mu_ow
```


**Contrast**

```{r}
get_contrast <- function(mu_tbl, g1, g2, method_name) {
  m1 <- mu_tbl$mean[mu_tbl$race == g1]
  m2 <- mu_tbl$mean[mu_tbl$race == g2]
  tibble(
    method = method_name,
    contrast = paste0(g1, " - ", g2),
    estimate = m1 - m2
  )
}

est_ipw <- bind_rows(
  get_contrast(mu_ipw, "White", "Black", "IPW"),
  get_contrast(mu_ipw, "White", "Asian", "IPW"),
  get_contrast(mu_ipw, "White", "Hispanic", "IPW")
)

est_ow <- bind_rows(
  get_contrast(mu_ow, "White", "Black", "Overlap"),
  get_contrast(mu_ow, "White", "Asian", "Overlap"),
  get_contrast(mu_ow, "White", "Hispanic", "Overlap")
)

bind_rows(est_ipw, est_ow)
```



**SE**

```{r}
estimate_pairwise_effects <- function(data) {
  # Refit multinomial GPS
  gps_fit_b <- nnet::multinom(gps_formula, data = data, trace = FALSE)
  gps_b <- as.matrix(predict(gps_fit_b, type = "probs"))
  
  data$gps_obs <- gps_b[cbind(seq_len(nrow(data)), match(data$race, colnames(gps_b)))]
  
  # IPW
  data$w_ipw <- 1 / data$gps_obs
  
  # Overlap
  h_b <- 1 / rowSums(1 / gps_b)
  data$w_ow <- h_b / data$gps_obs
  
  mu_ipw_b <- weighted_group_means(data, y = "healthExp", trt = "race", w = "w_ipw")
  mu_ow_b  <- weighted_group_means(data, y = "healthExp", trt = "race", w = "w_ow")
  
  c(
    ipw_white_black = mu_ipw_b$mean[mu_ipw_b$race == "White"] - mu_ipw_b$mean[mu_ipw_b$race == "Black"],
    ipw_white_asian = mu_ipw_b$mean[mu_ipw_b$race == "White"] - mu_ipw_b$mean[mu_ipw_b$race == "Asian"],
    ipw_white_hisp  = mu_ipw_b$mean[mu_ipw_b$race == "White"] - mu_ipw_b$mean[mu_ipw_b$race == "Hispanic"],
    ow_white_black  = mu_ow_b$mean[mu_ow_b$race == "White"] - mu_ow_b$mean[mu_ow_b$race == "Black"],
    ow_white_asian  = mu_ow_b$mean[mu_ow_b$race == "White"] - mu_ow_b$mean[mu_ow_b$race == "Asian"],
    ow_white_hisp   = mu_ow_b$mean[mu_ow_b$race == "White"] - mu_ow_b$mean[mu_ow_b$race == "Hispanic"]
  )
}
```


```{r}
set.seed(768)

B <- 500   
boot_mat <- replicate(B, {
  idx <- sample(seq_len(nrow(meps)), size = nrow(meps), replace = TRUE)
  dat_b <- meps[idx, ]
  estimate_pairwise_effects(dat_b)
})

boot_mat <- t(boot_mat)
boot_df <- as.data.frame(boot_mat)

apply(boot_df, 2, sd)
```



```{r}
point_est <- estimate_pairwise_effects(meps)

se_est <- apply(boot_df, 2, sd)

final_tbl <- tibble(
  term = names(point_est),
  estimate = as.numeric(point_est),
  se = as.numeric(se_est)
) %>%
  mutate(
    lwr = estimate - 1.96 * se,
    upr = estimate + 1.96 * se,
    method = case_when(
      str_detect(term, "^ipw") ~ "IPW",
      str_detect(term, "^ow")  ~ "Overlap"
    ),
    contrast = case_when(
      str_detect(term, "white_black") ~ "White - Black",
      str_detect(term, "white_asian") ~ "White - Asian",
      str_detect(term, "white_hisp")  ~ "White - Hispanic"
    )
  ) %>%
  dplyr::select(method, contrast, estimate, se, lwr, upr)

final_tbl


final_tbl %>%
  arrange(contrast, method)
```

